--- 
title: "The balanced ANOVA model with random effects"
author: "Stéphane Laurent"
date: '2017-07-29'
output:
  md_document:
    toc: yes
    variant: markdown
  html_document:
    keep_md: yes
prettify: yes
prettifycss: twitter-bootstrap
tags: maths, statistics
highlighter: kate
--- 


```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.path="./figures/AV1random-")
library(scales)
source('./Rfunctions/inertia_macro_v1.R', encoding='UTF-8')
```

\(\newcommand{\indic}{\mathbf{1}}\)
\(\newcommand{\perpoplus}{\overset{\perp}{\oplus}}\)
\(\newcommand{\RR}{\mathbb{R}}\)

## The balanced ANOVA model with random effects  

The balanced ANOVA model is used to model a sample $y=(y_{ij})$ with a tabular structure:
$$y=\begin{pmatrix}
y_{11} & \ldots & y_{1J} \\
\vdots & y_{ij} & \vdots \\
y_{I1} & \ldots & y_{IJ}
\end{pmatrix},
$$
$y_{ij}$ denoting the $j$-th response in group $i$. 

In the model with *random effects*, it is assumed that the observations $y_{ij}$ 
are generated in two steps. 
First, the group means $\mu_i$ are independently generated according to a 
Gaussian distribution ${\cal N}(\mu, \sigma^2_b)$ where $\mu$ is the overall 
mean and $\sigma^2_b$ is the so-called *between variance*. 
Second, the responses $y_{ij}$, $j =1,\ldots,J$, for each group $i$, are 
independently distributed according to  a Gaussian distribution 
${\cal N}(\mu_i, \sigma^2_w)$ with *within variance* $\sigma^2_w$ and mean
$\mu_i$. Shortly, the model can be written:
$$
\begin{cases}
\mu_i \sim_{\text{iid}} {\cal N}(\mu, \sigma^2_b) & i=1, \ldots, I \\
 (y_{ij} \mid \mu_i) \sim_{\text{iid}} {\cal N}(\mu_i, \sigma^2_w) 
 & j = 1, \ldots, J 
\end{cases}.
$$

```{r anovarandommodel, echo=FALSE, fig.width=9, fig.height=6}
set.seed(314) 
mu <- 24
delta <- 1.2
sigma <- 0.8
I <- 3
J <- 4
factor <- factor(rep(c("A","B","C"),rep(J,3)))
mu1 <- rnorm(1, mu, delta)
mu2 <- rnorm(1, mu, delta)
mu3 <- rnorm(1, mu, delta)
results <- c(
  rnorm(J, mu1, sigma), rnorm(J, mu2, sigma), rnorm(J, mu3, sigma)
)  
dat <- data.frame(factor=factor, result=results)
grandmean <- mean(dat$result)
omeans <- aggregate(result~factor, data=dat, FUN=mean)[,"result"]
#
fplot <- function(dat, results, ocolors, cols=rep("black",nrow(dat)), 
                  lcol="white", ylab, labs, cex=1.3, col.yaxis="black", 
                  cex.axis=1, cex.xlab=1.7, cex.ylab=1.7, padj=0){
  plot(c(17.5, 29.5), c(0.7, 5.8), type="n", axes=FALSE, 
       xlab="observation", ylab=NA, cex.lab=cex.xlab)
  axis(1, cex.axis=cex.axis, padj=padj)
  axis(2, labels=labs, at=c(3,2,1), lwd=0, cex.axis=1.7, col.axis = col.yaxis)
  mtext(ylab, 2, adj=0.23, padj=-3, cex=cex.ylab)
  h <- 4.5 # plateau du haut 
  abline(h=h)
  points(results, rep(h,length(results)), col=cols, pch=19, cex=cex)
  abline(h=1:3, col=lcol)
  k <- 1
  op <- levels(dat$factor)[k]
  x <- dat$result[dat$factor==op]
  points(x, rep(k,length(x)), col=ocolors[k], pch=19, cex=cex)
  k <- 2
  op <- levels(dat$factor)[k]
  x <- dat$result[dat$factor==op]
  points(x, rep(k,length(x)), col=ocolors[k], pch=19, cex=cex)
  k <- 3
  op <- levels(dat$factor)[k]
  x <- dat$result[dat$factor==op]
  points(x, rep(k,length(x)), col=ocolors[k], pch=19, cex=cex)
}
ocolors <- alpha(c("red","blue","green"),0.5)
cols <- ocolors[as.numeric(factor)]
h <- 4.5
cex.axis <- 1.8
ml <- 8
par(mar=c(5.9, ml, 0, 1))
K <- 1.8 # scale for gaussian curve
fplot(dat, results=omeans, ocolors=ocolors, cols=ocolors, lcol="black", 
      ylab=NA, labs=c("","group",""))
curve(K*dnorm(x,grandmean,delta)+h, add=TRUE, lwd=3, col=1)
text(grandmean, h-0.2, labels=expression(mu), cex=1.4, col=1) 
k <- 1
segments(omeans[k], h, omeans[k], k, col="red", lty=3)
k <- 2
segments(omeans[k], h, omeans[k], k, col="blue", lty=3)
k <- 3
segments(omeans[k], h, omeans[k], k, col="green", lty=3)
k <- 1
curve(K*dnorm(x,omeans[k],sigma)+h-1.5-(3-k), add=TRUE, lwd=2, col="red")
text(omeans[k], h-1.5-(3-k)-0.2, labels=expression(mu[1]), cex=1.4, col="red") 
k <- 2
curve(K*dnorm(x,omeans[k],sigma)+h-1.5-(3-k), add=TRUE, lwd=2, col="blue")
text(omeans[k], h-1.5-(3-k)-0.2, labels=expression(mu[2]), cex=1.4, col="blue")  
k <- 3
curve(K*dnorm(x,omeans[k],sigma)+h-1.5-(3-k), add=TRUE, lwd=2, col="green")
text(omeans[k], h-1.5-(3-k)-0.2, labels=expression(mu[3]), cex=1.4, col="green")
width <- shift <- 0.05
inertia(x0=grandmean, y0=h-0.35, a=1.6, b=0.15, lwd=2, l=0.25, col=c("green","blue","red"), w=width, s=shift)
a <- 1.1
b <- 0.11
vadj <- 0.3  
r <- 1/4
width <- shift <- 0.05
k <- 1
inertia(x0=omeans[k], y0=k-vadj, a=a, b=b, lwd=2, l=0.2, col="red", r=r, w=width, s=shift)
k <- 2
inertia(x0=omeans[k], y0=k-vadj, a=a, b=b, lwd=2, l=0.2, col="blue", r=r, w=width, s=shift)
k <- 3
inertia(x0=omeans[k], y0=k-vadj, a=a, b=b, lwd=2, l=0.2, col="green", r=r, w=width, s=shift)
```


An equivalent writing of this model, and from now on using capital letters for random 
variables, is 
$$Y_{ij} = \mu + \sigma_bA_{i} + \sigma_wG_{ij},$$
where all random variables $A_i$ and $G_{ij}$ are independent and $\sim {\cal N}(0,1)$. 



## The three summary statistics $\bar{Y}_{\bullet\bullet}$, $SS_b(Y)$, $SS_w(Y)$

Using the tensor product language introduced in [this article](http://stla.github.io/stlapblog/posts/Anova1fixed.html), the model can be written 
$$Y = \mu({\bf 1}_I\otimes{\bf 1}_J) + \sigma_b A\otimes\indic_J +\sigma_wG, 
\qquad A \sim SN(\RR^I), \quad G \sim SN(\RR^I\otimes\RR^J).$$


The overall mean $\bar{Y}_{\bullet\bullet}$ is given by the projection of $Y$ on the subspace 
$[{\bf 1}_I]\otimes[{\bf 1}_J]$:  
$$P_{[{\bf 1}_I]\otimes[{\bf 1}_J]} Y = \bar{Y}_{\bullet\bullet}({\bf 1}_I\otimes{\bf 1}_J).$$

Then the variations $Y_{ij}-\bar{Y}_{\bullet\bullet}$ around the overall mean are given by the projection on the orthogonal complement ${\Bigl([\indic_I]\otimes[\indic_J]\Bigr)}^\perp$. 
Knowing that 
$$
\RR^I \otimes \RR^J  = 
\Bigl([\indic_I]\otimes[\indic_J]\Bigr) 
\perpoplus \Bigl([\indic_I]^\perp\otimes[\indic_J]\Bigr) 
\perpoplus \Bigl([\indic_I]\otimes[\indic_J]^\perp\Bigr) 
\perpoplus \Bigl([\indic_I]^\perp\otimes[\indic_J]^\perp\Bigr),
$$
one gets
$$
\begin{align}
\underset{\text{'total'}}{\underbrace{ {\Bigl([\indic_I]\otimes[\indic_J]\Bigr)}^\perp} }  & = 
\Bigl([\indic_I]^\perp\otimes[\indic_J]\Bigr) 
\perpoplus \Bigl([\indic_I]\otimes[\indic_J]^\perp\Bigr) 
\perpoplus \Bigl([\indic_I]^\perp\otimes[\indic_J]^\perp\Bigr) \\
& =  \underset{\text{'between'}}{\underbrace{\Bigl([\indic_I]^\perp\otimes[\indic_J]\Bigr)}} 
\perpoplus 
\underset{\text{'within'}}{\underbrace{\Bigl(\RR^I\otimes[\indic_J]^\perp\Bigr)}},
\end{align}
$$
thereby yielding the decomposition of the variations:
$$
P^\perp_{[\indic_I]\otimes[\indic_J]}Y = 
P_{[\indic_I]^\perp\otimes[\indic_J]}Y + P_{\RR^I\otimes[\indic_J]^\perp}Y,
$$
whose component formulae are: 

- ${\bigl(P^\perp_{[\indic_I]\otimes[\indic_J]}Y\bigr)}_{ij}=Y_{ij}-\bar{Y}_{\bullet\bullet}$

- ${\bigl(P_{[\indic_I]^\perp\otimes[\indic_J]}Y \bigr)}_{ij} = \bar{Y}_{i\bullet}-\bar{Y}_{\bullet\bullet}$ 

- ${\bigl(P_{\RR^I\otimes[\indic_J]^\perp}Y\bigr)}_{ij} = Y_{ij}-\bar{Y}_{i\bullet}$ 

Now we can see that the three summary statistics (*overall mean*, *between sum of squares*, *within sum of squares*)
$$\bar{Y}_{\bullet\bullet}, \quad 
SS_b(Y):={\Vert P_{[\indic_I]^\perp\otimes[\indic_J]}Y  \Vert}^2, \quad 
SS_w(Y):={\Vert P_{\RR^I\otimes[\indic_J]^\perp}Y \Vert}^2,$$
are independent random variables.
Indeed, the overall mean $\bar{Y}_{\bullet\bullet}$ is given by 
$$\begin{align}
P_{[{\bf 1}_I]\otimes[{\bf 1}_J]} Y &= \bar{Y}_{\bullet\bullet}({\bf 1}_I\otimes{\bf 1}_J) \\
& =  
\mu({\bf 1}_I\otimes{\bf 1}_J) + \sigma_b(P_{[{\bf 1}_I]}A)\otimes\indic_J+\sigma_wP_{[{\bf 1}_I]\otimes[{\bf 1}_J]}G,
\end{align}$$
the between variations are 
$$P_{[\indic_I]^\perp\otimes[\indic_J]}Y 
= \sigma_b(P^\perp_{[\indic_I]} A)\otimes\indic_J + \sigma_w P_{[\indic_I]^\perp\otimes[\indic_J]} G,$$
and the within variations are 
$$P_{\RR^I\otimes[\indic_J]^\perp}Y 
= \sigma_wP_{\RR^I\otimes[\indic_J]^\perp} G.$$
Independence follows from the independence between $G$ and $A$ and 
from the orthogonality between the  ranges of the different projections.

It is easy to derive $\bar{Y}_{\bullet\bullet} \sim {\cal N}\left(\mu, \frac{\sigma^2}{IJ}\right)$ where $\sigma^2=J\sigma^2_b+\sigma^2_w$. 
It is also easy to get $SS_w(Y) \sim \sigma^2_w\chi^2_{I(J-1)}$ because of 
$$P_{\RR^I\otimes[\indic_J]^\perp}Y = \sigma_wP_{\RR^I\otimes[\indic_J]^\perp}G.$$ 
To derive the law of $SS_b(Y)$, 
note that 
$$
\begin{align}
P_{[\indic_I]^\perp\otimes[\indic_J]} G 
& = \begin{pmatrix} 
\bar{G}_{1\bullet} - \bar{G}_{\bullet\bullet} & \ldots & \bar{G}_{1\bullet} - \bar{G}_{\bullet\bullet} \\
\vdots & \vdots & \vdots \\
\bar{G}_{I\bullet} - \bar{G}_{\bullet\bullet} & \ldots & \bar{G}_{I\bullet} - \bar{G}_{\bullet\bullet} 
\end{pmatrix} \\
& = (P^\perp_{[\indic_I]}G_{\text{row}}) \otimes \indic_J
\end{align}$$
where $G_{\text{row}} = (\bar{G}_{i\bullet})$ is the vector of row means of $G$, and then one can write  
$$P_{[\indic_I]^\perp\otimes[\indic_J]} Y 
= \bigl(P^\perp_{[\indic_I]}(\sigma_b A + \sigma_w G_{\text{row}})\bigr) \otimes \indic_J.$$
Now it is easy to see that the components of $\sigma_b A + \sigma_w G_{\text{row}}$ are 
$\sim_{\text{iid}} {\cal N}(0, \sigma^2)$, and 
consequently $SS_b(Y) \sim \sigma^2\chi^2_{J-1}$. 


## Confidence interval for the overall mean 

By our previous derivations, the statistic 
$$ \frac{\bar Y_{\bullet\bullet}  - \mu}{\frac{1}{\sqrt{I}}\sqrt{\frac{SS_b(Y)}{I(J-1)}}}$$
has a Student $t$ distribution with $I-1$ degrees of freedom, wherefrom it is easy to get an exact confidence interval about the overall mean $\mu$. 

Note that we would get exactly the same confidence interval if we were told only the group means $\bar{Y}_{i\bullet}$. 
 This is the topic of [another article](http://stla.github.io/stlapblog/posts/ModelReduction.html).


