---
title: "The binary splitting with Haskell"
author: "Stéphane Laurent"
highlighter: hscolour
date: "2017-06-05"
tags: haskell, special-functions
output:
  md_document:
    variant: markdown
  html_document: default
prettify: yes
---


```{r setup, include=FALSE}
knitr::knit_engines$set(ghc = function (options)
{
  engine = options$engine
  f = basename(tempfile(engine, ".", ".txt"))
  writeLines(options$code, f)
  on.exit(unlink(f))
  code = paste(f, options$engine.opts)
  cmd = options$engine.path
  out = if (options$eval) {
    message("running: ", cmd, " ", code)
    tryCatch(system2(cmd, code, stdout = TRUE, stderr = FALSE,
                     env = options$engine.env), error = function(e) {
                       if (!options$error)
                         stop(e)
                       paste("Error in running command", cmd)
                     })
  }
  else ""
  if (!options$error && !is.null(attr(out, "status")))
    stop(paste(out, collapse = "\n"))
  knitr::engine_output(options, options$code, out)
}
)
## chunk options
knitr::opts_chunk$set(engine = 'ghc', 
                      engine.path = 'ghcscriptrender', 
                      engine.opts = '--fragment --singleoutputs', 
                      echo = FALSE, results = 'asis')
```

At the first line of each script in this article, we'll load the following small Haskell module:

```{r, engine.opts="--fragment --module"}
-- BinarySplitting.hs
module BinarySplitting
  where
import Data.Ratio ((%))

split0 :: ([Rational], [Rational]) -> [Rational]
split0 u_v = map (\i -> (u !! (2*i)) * (v !! (2*i+1))) [0 .. m]
  where (u, v) = u_v
        m = div (length u) 2 - 1

split1 :: ([Rational], [Rational], [Rational]) ->
               ([Rational], [Rational], [Rational])
split1 adb = split adb (length alpha)
  where (alpha, _, _) = adb
        split :: ([Rational], [Rational], [Rational]) -> Int ->
                             ([Rational], [Rational], [Rational])
        split u_v_w n =
          if n == 1
            then u_v_w
            else split (x, split0 (v,v), split0 (w,w)) (div n 2)
          where (u, v, w) = u_v_w
                x  = zipWith (+) (split0 (u, w)) (split0 (v, u))

bsplitting :: Int -> [Rational] -> [Rational] -> Rational
bsplitting m u v = num / den + 1
  where ([num], _, [den]) = split1 (take (2^m) u, take (2^m) u, take (2^m) v)
```

The `bsplitting` function performs the [binary splitting algorithm](https://laustep.github.io/stlahblog/posts/hypergeometric.html). 

Given an integer $m \geq 0$ and two sequences $(u_i)$ and $(v_i)$ 
of rational numbers, it calculates the sum
$$
A_m = 1 + \sum_{k=1}^{2^m} \prod_{i=1}^k\frac{u_i}{v_i}.  
$$

## Approximation of $\pi$

For example, $A_m \to \frac{\pi}{2}$ when $u_i = i$ and $v_i = 2i+1$. 
So we get a rational approximate of $\pi$ as follows:

```{r}
:load BinarySplitting.hs

:{
approxPi :: Int -> Rational
approxPi m = 2 * bsplitting m u v
  where u = [i | i <- [1 ..]]
        v = [2*i+1 | i <- [1 ..]]
:}

let x = approxPi 5
x
fromRational x
```


## Kummer hypergeometric function

Consider the confluent hypergeometric series 
$$
{}_1\!F_1(a, b; x) = \sum_{n=0}^{\infty}\frac{{(a)}_{n}}{{(b)}_{n}}\frac{x^n}{n!} = 1 + \sum_{n=1}^{\infty}\frac{{(a)}_{n}}{{(b)}_{n}}\frac{x^n}{n!}.
$$
 Here ${(a)}_n:=a(a+1)\cdots(a+n-1)$ is the Pocchammer symbol denoting the ascending factorial.
 The sum from $n=0$ to $n=2^m$ is evaluated by the `bsplitting` function by taking the sequences  
$u_i = (a+i-1)x$ and $v_i = (b+i-1)i$.

Below we evaluate this sum for $a=8.1$, $b=10.1$ and $x=100$. 
We compare the result to the value of ${}_1\!F_1(a, b; x)$ given by Wolfram.

```{r}
:load BinarySplitting.hs

:{
hypergeo1F1 :: Int -> Rational -> Rational -> Rational -> Double
hypergeo1F1 m a b x = fromRational $ bsplitting m u v
  where u = [(a+i)*x | i <- [0 ..]]
        v = [(b+i)*(i+1) | i <- [0 ..]]
:}

let wolfram = 1.7241310759926883216143646e41

wolfram - hypergeo1F1 6 (81%10) (101%10) 100
wolfram - hypergeo1F1 7 (81%10) (101%10) 100
wolfram - hypergeo1F1 8 (81%10) (101%10) 100
```

We find a good approximate for $m=8$ (so $2^m=256$), and the evaluation is really fast.