---
title: "The conditional law as a random probability"
author: "Stéphane Laurent"
date: '2020-05-10'
tags: maths
output:
  md_document:
    variant: markdown
    preserve_yaml: true
  html_document:
    highlight: kate
    keep_md: no
prettifycss: minimal
highlighter: pandoc-solarized
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.path = "figures/conditionalLaw-")
```

Let $(\Omega, \mathfrak{A}, \mathbb{P})$ be a probability space, 
$E$ a Polish space, and $X$ an $E$-valued random variable. 
Denote by $\mathcal{B}_E$ the Borel $\sigma$-field on $E$. 
It is well-known that the law of $X$ can be disintegrated over 
any $\sigma$-field $\mathfrak{B}\subset\mathfrak{A}$. 
That means that there exists a *probability kernel* 
$K\colon\Omega\times\mathcal{B}_E \to [0,1]$ from $\Omega$ to $E$ 
such that for every suitable function $f\colon E \to\mathbb{R}$,
$$
\mathbb{E}\bigl[f(X)\bigr] = 
\int_\Omega K_\omega(f) \mathrm{d}\mathbb{P}(\omega)
$$
where we denote by $K_\omega$ the probability $A\mapsto K(\omega,A)$ 
and we set $\mu(f) = \int f \mathrm{d}\mu$ for a probability $\mu$. 
The map $\omega\mapsto K_\omega$ is called the 
*conditional law of $X$ given $\mathfrak{B}$*, and is denoted by 
$\mathcal{L}(X\mid\mathfrak{B})$.

Recall that a *probability kernel* from a probability space $\Omega$ 
to a measurable space $(E,\mathcal{B})$ is 
an application $K\colon\Omega\times\mathcal{B} \to [0,1]$ such that

*(i)* for fixed $\omega\in\Omega$, the map $A \mapsto K(\omega,A)$ is a 
probability on $(E,\mathcal{B})$;

*(ii)* for fixed $A\in\mathcal{B}$, the map $\omega \mapsto K(\omega,A)$ 
is measurable.

It is less known that there exists a Polish topology on the set $\Pr(E)$ of 
probabilities on $E$ such that the conditional law 
$\mathcal{L}(X\mid\mathfrak{B})$ is a $\Pr(E)$-valued random variable. 

I am unable to find a self-contained reference for this result. 
I aim to prove it in these notes in a self-contained way, except 
that I don't give a proof of Prohorov's theorem here. 
However, I give a proof of Prohorov's theorem for Lusin spaces 
in another post ([Probabilities on a Lusin space](ProhorovLusin.html)), 
and Polish spaces are Lusin. 


# Polish spaces

A topological space $E$ is said to be *Lindelöf* if there is a 
countable subcover of every open cover of $E$. 
It is said to be *second-countable* when it has a countable basis. 

**Theorem.** *Every second-countable topological space is Lindelöf.*

*Proof.* Let $E$ be a second-countable topological space. 
Take a countable basis ${\{O_n\}}_{n \in \mathbb{N}}$ of the topology on $E$. 
Let $\mathcal{C}={\{C_i\}}_{i \in I}$ be an open cover of $E$. 
For every $x \in E$, take $i(x) \in I$ such that $x \in C_{i(x)}$, 
and let $n(x) \in \mathbb{N}$ such that $x \in O_{n(x)} \subset C_{i(x)}$. 
Then ${\{O_{n(x)}\}}_{x \in X}$ is a countable open cover of $E$. 
For every $\ell \in n(E) \subset \mathbb{N}$, the set 
$\{j \in I \mid O_\ell \subset C_j\}$ is non-empty, therefore one can 
select $j(\ell) \in I$ such that $O_\ell \subset C_{j(\ell)}$ by using 
the axiom of countable choice. 
The set ${\{C_{j(\ell)}\}}_{\ell \in n(E)}$ is a countable subcover of 
$\mathcal{C}$. &square;


**Theorem.** *For a metrizable space, the notions of separability, 
second-countable and Lindelöf are equivalent.*

*Proof.* Let $E$ be a metric space. Denote by $B_r(x)$ the open ball of 
radius $r$ centered at $x$. 

Assume $E$ is separable. Let 
${(x_n)}_{n\geqslant 1}$ be a dense sequence of $E$. Then 
$\bigl\{B_r(x_n) \mid r \in \mathbb{Q}, n\geqslant 1\bigr\}$ 
is a countable basis of $E$, hence $E$ is second-countable. 

By the previous theorem, the second-countable property implies the 
Lindelöf property. Now assume $E$ is Lindelöf. Consider the open cover 
$\bigl\{B_\epsilon(x)\bigr\}_{x \in E}$ for a given $\epsilon>0$. 
By the Lindelöf property, there exists a countable subcover 
$\bigl\{B_\epsilon(x_{\epsilon,k})\bigr\}_{k \in \mathbb{N}}$. 
Then $\bigcup_{n\geqslant 1} {\bigl\{x_{1/n,k}\bigr\}}_{k \in \mathbb{N}}$ 
is dense in $E$, thereby showing that $E$ is separable. 
&square;


**Corollary.** *A subspace of a metrizable separable space is itself separable*. 

*Proof.* It is clear that a subspace of a second-countable space is 
itself second-countable. Therefore the result follows from the previous 
theorem. &square;


Given a metric space $(E,d)$, a non-empty set $A \subset E$ and $\epsilon>0$, 
we define the set
$$
A^\epsilon = \bigl\{x \in E \mid d(x,A) < \epsilon\bigr\}.
$$
It is well-known that $x \mapsto d(x,A)$ is continuous. 
This straigthforwardly results from the inequality 
$\bigl|d(x,A) - d(y,A)\bigr| \leqslant d(x,y)$ (see [2]). 
Hence $A^\epsilon$ is open. 
It is also well-known that $d(x,A) = 0$ if and only if $x \in \overline{A}$. 


**Proposition.** *An open subset of a Polish space is Polish.* 

*Proof.* Let $E$ be a Polish metric space, and $A \subset E$ be open. 
Then $A$ is separable by the previous corollary. 
Assume $A \neq E$ and let $d$ be a compatible metric on $E$ under which 
$E$ is complete. For $x,y\in A$, define
$$
d'(x,y) = d(x,y) + \left|\frac{1}{d(x,A^c)}-\frac{1}{d(y,A^c)}\right|.
$$
It is easy to check that $d'$ is a metric on $A$. 
Using the continuity of $x \mapsto d(x,A^c)$, it is easy to see that 
convergence for $d$ is equivalent to convergence for $d'$. 

It is also clear that a $d'$-Cauchy sequence $(u_n)$ in $A$ is a 
$d$-Cauchy sequence. Let $u_\infty$ be the limit of $(u_n)$. 
Then $u_\infty \in A$, otherwise $d(u_n, A^c) \to 0$ and 
$d'(u_n,u_m) \to \infty$ as $m,n \to \infty$
&square;


More generally, a subset of a Polish space is a Polish space if and only if 
it is a $G_\delta$ set - see [6].

A topological space is said to be *strongly Lindelöf* if every 
open set is Lindelöf.

**Corollary.** *A Polish space is strongly Lindelöf.*

*Proof.* A Polish space is metrizable and separable, therefore it is Lindelöf 
by the second theorem. 
Since an open subset of a Polish space is Polish, we see that 
a Polish space is strongly Lindelöf. &square;


# Space of probability measures

For a topological space $E$, we denote by $\mathcal{B}_E$ 
the Borel $\sigma$-field on $E$, by $C_b(E)$ the space of 
real-valued bounded continuous functions on $E$ and by 
$\Pr(E)$ the topological space of probabilities on $E$ equipped with the 
narrow topology. 

The *canonical basis of neighborhoods* at $\mu \in \Pr(E)$ is the 
family of open sets
$$
\Bigl\{\nu\in\Pr(E) \mid
\bigl|\nu(f_i)-\mu(f_i)\bigl| < \epsilon, i\in [\![1,k]\!]\Bigr\}
$$
for $\epsilon>0$, $k \geqslant 1$ and $f_i \in C_b(E)$. 

In the last section of these notes, we will use the fact that 
a net $(\mu_\lambda)$ in $\Pr(E)$ converges to $\mu_\infty \in \Pr(E)$ 
if and only if $\mu_\lambda(f) \to \mu_\infty(f)$ for every $f \in C_b(E)$. 


**Lemma**. *Let $\mu\in\Pr(E)$, $f \in C_b(E)$ and $\epsilon > 0$. 
There exist an integer $k \geqslant 1$, a number $\eta > 0$ and 
some closed sets $F_1 \subset E, \ldots, F_k \subset E$ such that the 
set* 
$$
\Bigl\{\nu \mid  \bigl|\nu(f)-\mu(f)\bigl| < \epsilon\Bigr\}
$$
*contains the set*
$$
\bigl\{\nu \in \Pr(E) \mid 
\nu(F_i)<\mu(F_i)+\eta, i\in [\![1,k]\!]\bigr\}.
$$

*Proof.* Using a linear transformation, one can write
$$
\Bigl\{\nu \mid  \bigl|\nu(f)-\mu(f)\bigl| < \epsilon\Bigr\}
$$
as 
$$
\Bigl\{\nu \mid  \bigl|\nu(g)-\mu(g)\bigl| < \epsilon'\Bigr\}
$$
where $g \in C_b(E)$ is such that $0 < g(x) < 1$ for all $x\in E$ 
and $\epsilon' > 0$. 
Take an integer $k > \frac{2}{\epsilon'}$ and take the closed sets 
$F_i = \bigl\{x \in E \mid g(x) \geqslant \frac{i-1}{k}\bigr\}$ 
for $i\in [\![1,k]\!]$ and $F_{k+1} = \varnothing$.

Define the intervals $J_i = \bigl[\frac{i-1}{k}, \frac{i}{k} \bigr[$. 
For every $P\in\Pr(E)$, one has
$$
\sum_{i=1}^k \frac{i-1}{k}P\bigl(g^{-1}(J_i)\bigr) \leqslant P(g) \leqslant 
\sum_{i=1}^k \frac{i}{k}P\bigl(g^{-1}(J_i)\bigr).
$$
The sum on the right is
$$
\sum_{i=1}^k \frac{i}{k}\bigl(P(F_i) - P(F_{i+1})\bigr) 
= \frac{1}{k} + \frac{1}{k}\sum_{i=2}^kP(F_i)
$$
and the sum on the left is $\frac{1}{k}\sum_{i=2}^kP(F_i)$. 
Therefore $\nu(g) < \frac{\epsilon'}{2} + \frac{1}{k}\sum_{i=2}^k\nu(F_i)$ 
and $\frac{1}{k}\sum_{i=2}^k\mu(F_i) \leqslant \mu(g)$. 
Thus, if $\nu(F_i)<\mu(F_i)+\eta$, one has
$$
\nu(g) < \frac{\epsilon'}{2} + \mu(g) + \eta
$$
and then $\nu(g) < \mu(g) + \epsilon'$ if we take 
$\eta = \frac{\epsilon'}{2}$. 
Applying the same mathematics with $1-g$ instead of $g$, we finally get 
$\bigl|\nu(g)-\mu(g)\bigl| < \epsilon'$. &square;


The following proposition follows from this lemma; 
it will be used in [Probabilities on a Lusin space](ProhorovLusin.html).

**Proposition.** *Let $\mu\in\Pr(E)$ and $(\mu_n)$ be a sequence in $\Pr(E)$. 
If $\limsup \mu_n(F) \leqslant \mu(F)$ for all closed sets $F\subset E$, 
then $(\mu_n)$ converges to $\mu$.*

*Proof.*
Let $V$ be a neighborhood of $\mu$. By the previous lemma, there exist 
an integer $k \geqslant 1$, a number $\epsilon > 0$ and 
some closed sets $F_1 \subset E, \ldots, F_k \subset E$ such that 
$$
U := \bigl\{\nu \in \Pr(E) \mid 
\nu(F_i)<\mu(F_i)+\epsilon, i\in [\![1,k]\!]\bigr\} \subset V.
$$
One has $\limsup \mu_n(F_i) \leqslant \mu(F_i)$ for every $i\in [\![1,k]\!]$. 
Therefore, there is an integer $N_i$ such that 
$\mu_n(F_i) < \mu(F_i) + \epsilon$ when $n \geqslant N_i$. 
Hence, $\mu_n \in U$ for $n\geqslant\max\{N_i\}$.
Since $U\subset V$, one gets $\mu_n \to \mu$ in $\Pr(E)$.
&square;


**Theorem.** *The probabilities on $E$ with finite support form a dense 
subset of $\Pr(E)$.* 

*Proof.* Let $\mu \in \Pr(E)$ and 
$$
V = 
\Bigl\{\nu \in\Pr(E) \mid \bigl|\nu(f_i)-\mu(f_i)\bigl| < \epsilon, 
i\in [\![1,k]\!]\Bigr\}
$$
be an open set in the canonical basis of neighborhoods of $\mu$. 
By the previous lemma, $V$ contains a set 
$$
U = \bigl\{\nu \in \Pr(E) \mid 
\nu(F_i)<\mu(F_i)+\epsilon', i\in [\![1,k']\!]\bigr\}
$$
with $k'\geqslant 1$, $\epsilon'>0$, and $F_i\subset E$ closed. 
The $F_i$ generate a finite partition  
$\mathcal{P} = \{B_1, \ldots, B_J\}$ with $B_j \neq \varnothing$ closed. 
For each $B_j \in \mathcal{P}$, we pick a point  
$b_j \in B_j$. The probability $\sum_{j=1}^J\mu(B_j)\delta_{b_j}$ 
belongs to $U$. That shows the result. 
&square;


**Theorem.** *If $E$ is separable, then so is $\Pr(E)$.*

*Proof.* Let $D \subset E$ be a countable dense subset of $E$. 
We are going to show that the set of probability measures with 
finite support contained in $D$ and
which have a rational mass at each point of their support, is dense in 
$\Pr(E)$. Since this set is countable, this will prove the proposition. 

By the preceding theorem, we know that the probabilities with 
finite support form a dense subset of $\Pr(E)$. 
Let $\mu = \sum_{j=1}^J p_j \delta_{x_j}$ be such a probability. 
Take an open neighborhood of $\mu$ 
$$
\mathcal{V} = 
\Bigl\{\nu \in\Pr(E) \mid  \bigl|\nu(f_i)-\mu(f_i)\bigl| < \epsilon, 
i\in [\![1,k]\!]\Bigr\},
$$
with $k \geqslant 1$, $\epsilon>0$ and $f_i \in C_b(E)$. 
Note that $\mu(f_i) = \sum_{j=1}^J p_j f_i(x_j)$. 

Let $K = \max\bigl\{{\Vert f_i \Vert}_\infty\bigr\}$. 
Let $q_j \in \mathbb{Q}\cap [0,1]$, $j \in [\![1,J]\!]$, 
such that $\sum_{j=1}^J q_j = 1$ and $|q_j - p_j| < \frac{\epsilon}{2JK}$. 
Let $y_{j} \in D$ such that 
$\bigl|f_i(y_{j}) - f_i(x_j)\bigr| < \frac{\epsilon}{2}$ 
for every $j \in [\![1,J]\!]$.  

We set $\nu = \sum_{j=1}^J q_j \delta_{y_{j}}$, so that one has 
$\nu(f_i) = \sum_{j=1}^J q_j f_i(y_{j})$, and then 
$$
\bigl|\nu(f_i)-\mu(f_i)\bigr| \leqslant 
\sum_{j=1}^J \bigl|q_j f_i(y_{j}) - p_j f_i(x_j)\bigr|.
$$
Let us bound from above each term of the sum:
$$
\begin{align}
& \bigl|q_j f_i(y_{j}) - p_j f_i(x_j)\bigr|  \\ & \quad \leqslant 
\bigl|q_j f_i(y_{j}) - p_j f_i(y_{j})\bigr| + 
\bigl|p_j f_i(y_{j}) - p_j f_i(x_j)\bigr| \\ & \quad \leqslant
|q_j - p_j| {\Vert f_i \Vert}_\infty + p_j \bigl|f_i(y_{j}) - f_i(x_j)\bigr| \\ 
& \quad < \frac{\epsilon}{2J} + p_j \frac{\epsilon}{2}.
\end{align}
$$
By summing, one gets
$\bigl|\nu(f_i)-\mu(f_i)\bigr| < \epsilon$, hence $\nu \in \mathcal{V}$. 
&square;


**Theorem.** *Assume $E$ is metric and denote by $d$ the metric on $E$. 
Let $F \subset E$ be closed and $\epsilon>0$. Then 
there is a function $f\in C_b(E)$ such that $f(x)=1$ for $x\in F$, 
$f(x)=0$ if $d(x,F)\geqslant\epsilon$ and $0\leqslant f(x)\leqslant 1$ for all $x \in E$. 
This function $f$ may be taken to be uniformly continuous.* 

*Proof.* Define the continuous function $\phi\colon\mathbb{R}\to [0,1]$ by
$$
\phi(t) = 
\begin{cases}
1 & \text{if } t \leqslant 0 \\
1-t & \text{if } 0 \leqslant t \leqslant 1 \\
0 & \text{if } t \geqslant 1 
\end{cases}.
$$
Then define 
$$
f(x) = \phi\left(\frac{1}{\epsilon}d(x,F)\right).
$$
This function $f$ has the required properties. 
Here is the graph of $f$ for $F = [a,b] \subset \mathbb{R}$:

```{r graph, echo=FALSE, fig.width=7, fig.height=2}
par(mar = c(2,0,0,0))
plot(NULL, xlim = c(-5, 5), ylim = c(0,1), xlab = NA, ylab = NA, 
     axes = FALSE, yaxs = "i")
abline(h = 0)
segments(-4, 0, -3, 1, lwd = 3)
segments(-3, 1, 3, 1, lwd = 3)
segments(3, 1, 4, 0, lwd = 3)
axis(1, labels = c(expression(italic(a)-epsilon), expression(italic(a)), 
                   expression(italic(b)), expression(italic(b)+epsilon)), 
     at = c(-4,-3,3,4), lwd = 0, lwd.ticks = 1, cex.axis = 1.2)
```

&square;


**Proposition (neighborhood bases of the narrow topology).** *If $E$ is a 
metric space, then each of the three families 
of sets below form a neighborhood basis of the narrow topology at 
$\mu \in \Pr(E)$.*

*(i)* $\bigl\{\nu \in \Pr(E) \mid \nu(F_i)<\mu(F_i)+\epsilon, i\in [\![1,k]\!]\bigr\}$, 
*where $k \geqslant 1$ is an integer, the $F_i$ are closed in $E$, and* 
$\epsilon > 0$.

*(ii)* $\bigl\{\nu \in \Pr(E) \mid \nu(O_i)>\mu(O_i)-\epsilon, i\in [\![1,k]\!]\bigr\}$, 
*where $k \geqslant 1$ is an integer, the $O_i$ are open in $E$, and* 
$\epsilon > 0$.

*(iii)* $\Bigl\{\nu \in \Pr(E) \mid \bigl|\nu(A_i)-\mu(A_i)\bigr|<\epsilon, i\in [\![1,k]\!]\Bigr\}$, 
*where $k \geqslant 1$ is an integer, the $A_i$ are Borelian in $E$ and satisfy* 
$\mu(\partial A_i)=0$, *and* $\epsilon > 0$.

*Proof.* One obviously gets equality between the families *(i)* and *(ii)* 
with the help of the complementary operation. 
By the previous lemma, each neighborhood at $\mu$ in the canonical 
basis contains a set in the family *(ii)*. Let us show the converse. 
Denote by $d$ the metric on $E$ and for $A \subset E$ and $\eta > 0$, 
define $A^\eta = \bigl\{x\in E \mid d(x,A) < \eta \bigr\}$. 
Choose $\eta_i>0$ such that 
$\mu(F_i^{\eta_i}) < \mu(F_i) + \frac{\epsilon}{2}$. 
Apply the previous theorem to get $f_i\in C_b(E)$ with value $1$ on $F_i$, 
value $0$ outside $F_i^{\eta_i}$, and value everywhere contained in $[0,1]$. 
If $\bigl|\nu(f_i)-\mu(f_i)\bigr| < \frac{\epsilon}{2}$, then 
$$
\nu(F_i) \leqslant \nu(f_i) < \mu(f_i) + \frac{\epsilon}{2} 
\leqslant \mu(F_i^{\eta_i}) + \frac{\epsilon}{2} < \mu(F_i) + \epsilon.
$$
Thus, each set in the canonical neighborhood at $\mu$ is contained in 
a set of family *(i)*. 

It remains to treat the family *(iii)*. If $A\subset E$ is Borel and 
$\mu(\partial A) = 0$, then for $\nu$ in a set of family *(i)* we have 
$$
\nu(A) \leqslant \nu(\overline{A}) < \mu(\overline{A}) + \epsilon 
= \mu(A) + \epsilon
$$
and for $\nu$ in a set of family *(ii)* we have 
$$
\nu(A) \geqslant \nu(A^\circ) > \mu(A^\circ) - \epsilon 
= \mu(A) - \epsilon.
$$
Since *(i)* is a neighborhood basis at $\mu$, each set of family *(iii)* 
contains a set of family *(i)*. 
On the other hand, if $F \subset E$ is closed, then for any $\epsilon>0$ 
there is $\eta>0$ such that $\mu(\partial F^\eta) = 0$ and 
$\mu(F^\eta) < \mu(F) + \frac{\epsilon}{2}$. 
Indeed, firstly one can obviously take $\lambda$ such that 
$\mu(F^\lambda) < \mu(F) + \frac{\epsilon}{2}$. Since the 
$\partial F^\eta$ for $\eta \in ]0, \lambda[$ have empty intersection 
(since $\partial F^\eta = \bigl\{x\in E \mid d(x,F) = \eta \bigr\}$), 
there is $\eta \in ]0, \lambda[$ such that $\mu(\partial F^\eta) = 0$. 
Obviously $\mu(F^\eta) \leqslant \mu(F^\lambda)$. The claim is proved. 
Now take $\nu$ such that 
$\bigl|\nu(F^\eta)-\mu(F^\eta)\bigr| < \frac{\epsilon}{2}$. Then
$$
\nu(F) \leqslant \nu(F^\eta) < \mu(F^\eta) + \frac{\epsilon}{2} 
< \mu(F) + \epsilon.
$$
Thus each set of family *(i)* contains a set of family *(iii)*. 
&square;


The following proposition will be used in 
[Probabilities on a Lusin space](ProhorovLusin.html). 

**Proposition.** *Assume $E$ is metric. 
Let $\mu\in\Pr(E)$ and $(\mu_n)$ be a sequence in $\Pr(E)$, 
such that $(\mu_n)$ converges to $\mu$. Then 
$\limsup \mu_n(F) \leqslant \mu(F)$ for all closed sets $F\subset E$.*

*Proof.* Suppose there exists a closed $F\subset E$ such that 
$\limsup \mu_n(F) > \mu(F)$. Let $\epsilon>0$ such that 
$\limsup \mu_n(F) > \mu(F)+\epsilon$. 
Then there are infinitely many $n$ 
such that $\mu_n(F) \geqslant \mu(F)+\epsilon$. 
The set 
$$
V = \bigl\{\nu \in \Pr(E) \mid \nu(F)<\mu(F)+\epsilon\bigr\}
$$
is a neighborhood of $\mu$ by the previous proposition. 
But for every $N \geq 0$, there exists $n \geqslant N$ for which 
$\mu_n\not\in V$. That shows that $(\mu_n)$ does not converge to $\mu$. 
&square;


**Lemma (regularity of probabilities on metric spaces)** *Assume $E$ is metric. 
Then for every probability $\mu$ on $E$ and every Borel set $A \subset E$,*
$$
\begin{align}
\mu(A) & = \inf\bigl\{\mu(O) \mid O \text{ open}, O \supset A\bigr\} \\
& = \sup\bigl\{\mu(F) \mid F \text{ closed}, F \subset A\bigr\}.
\end{align}
$$

*Proof.* The first equality is obviously true for an open set $A$. 
Let us show that the second equality is true for any open set $A$. 
There exist some closed sets $F_n \uparrow A$: take 
$F_n = \bigl\{x \mid d(x,A^c) \geqslant \frac{1}{n} \bigr\}$. 
Since we have $\mu(F_n) \uparrow \mu(A)$, it is easy to see that the second 
equality holds for $A$. Let $\mathcal{A}$ be the set of all Borel sets $A$ 
fulfilling the stated property. To prove that the lemma is true, it 
remains to show that $\mathcal{A}$ is a $\sigma$-algebra. 
Let us show this is the case. It is clear that $\mathcal{A}$ is stable 
under complementation. 
It is easy to see that $A \in \mathcal{A}$ if and only if for every 
$\epsilon > 0$ there exist an open set $O \supset A$ and a closed set 
$F \subset A$ such that $\mu(O\setminus F) < \epsilon$.

Let $(A_n)$ be a sequence in $\mathcal{A}$. 
For every $n$, let $F_n \subset A_n \subset O_n$ with 
$\mu(O_n \setminus F_n) \leqslant \epsilon/3^n$. 
Let $n_0$ such that 
$$
\mu\bigl(\bigcup F_n \setminus \bigcup_{k=1}^{n_0} F_k \bigr) < 
\frac{\epsilon}{2}.
$$
Take $O = \bigcup_n O_n$ and $F = \bigcup_{k=1}^{n_0} F_k$. 
Then $F \subset \bigcup_n A_n \subset O$ and since 
$O \setminus \bigcup_n F_n \subset \bigcup (O_n\setminus F_n)$, 
$$
\mu(O\setminus F) \leqslant 
\mu(O \setminus \bigcup F_n) + \mu(\bigcup F_n \setminus F) < 
\sum \mu(O_n\setminus F_n) + \frac{\epsilon}{2} < \epsilon.
$$
&square;


**Lemma.** *If the inequality $\mu(A) \leqslant \nu(A^\epsilon) + \epsilon$ 
holds for $A = E\setminus B^\epsilon$, then 
$\nu(B) \leqslant \mu(B^\epsilon) + \epsilon$.*

*Proof.* Note that the two inclusions $A \subset E\setminus B^\epsilon$ and 
$B \subset E\setminus A^\epsilon$ are equivalent because each one is 
equivalent to $d(x,y) \geqslant \epsilon$ for all $x\in A$ and $y\in B$. 
Hence
$$
\mu(B^\epsilon) = 1 - \mu(A) \geqslant 1 - \nu(A^\epsilon)-\epsilon 
= \nu(E\setminus A^\epsilon)-\epsilon \geqslant \nu(B) -\epsilon.
$$
&square;

We will not use equality $(2)$ of the following proposition.

**Proposition and definition (Prohorov distance).** *Let $E$ be a metric 
space. 
The Prohorov distance 
between $\mu \in \Pr(E)$ and $\nu \in \Pr(E)$ is defined by*
$$
d_P(\mu,\nu) = 
\inf\bigl\{\epsilon>0 \mid \forall A\in\mathcal{B}_E, 
\mu(A) \leqslant \nu(A^\epsilon) + \epsilon \,\text{ and }\, 
\nu(A) \leqslant \mu(A^\epsilon) + \epsilon\bigr\}.
$$
*One has* 
$$
\begin{align}
d_P(\mu,\nu) & \overset{(1)}{=} 
\inf\bigl\{\epsilon>0 \mid \forall A\in\mathcal{B}_E, 
\mu(A) \leqslant \nu(A^\epsilon) + \epsilon\bigr\} \\ & \overset{(2)}{=}
\inf\bigl\{\epsilon>0 \mid \forall \text{ closed } F\subset E, 
\mu(F) \leqslant \nu(F^\epsilon)+\epsilon\bigr\}.
\end{align}
$$

*Proof.* Let $\epsilon>0$ such that 
$\mu(A) \leqslant \nu(A^\epsilon) + \epsilon$ for all $A\in\mathcal{B}_E$. 
Let $B\in\mathcal{B}_E$. Then $E\setminus B^\epsilon\in\mathcal{B}_E$ 
since $B^\epsilon$ is open. 
Applying the previous lemma, we get 
$\nu(B) \leqslant \mu(B^\epsilon) + \epsilon$. 
Thus the two inequalities in the definition of $d_P(\mu,\nu)$ hold 
for all $B\in\mathcal{B}_E$. Equality $(1)$ is proved.

Now we prove equality $(2)$. 
Let $\epsilon>0$ such that $\mu(F) \leqslant \nu(F^\epsilon)+\epsilon$ 
for all closed sets $F \subset E$. 
We will first prove that $\nu(F) \leqslant \mu(F^\epsilon)+\epsilon$ 
for all closed sets $F \subset E$. Let $H\subset E$ be closed and 
$G = E\setminus H^\epsilon$. Then $G$ is closed. Hence 
$\nu(H) \leqslant \mu(H^\epsilon) + \epsilon$ by the previous lemma.
Thus it suffices to prove that $\epsilon>0$ satisfies
$$
\mu(A) \leqslant \nu(A^\epsilon) + \epsilon \,\text{ and }\, 
\nu(A) \leqslant \mu(A^\epsilon) + \epsilon
$$
for all Borel $A$ if and only if it satisfies these inequalities for 
all closed $A$. 
Assume these inequalities hold for all closed $A$. Let $B\subset E$ be Borel. 
Take $\eta>0$. By the lemma *"regularity of probabilities on metric spaces"*, 
there exists a closed $C\subset B$ 
such that $\mu(C) \geqslant \mu(B)-\delta$. Hence
$$
\mu(B)-\delta \leqslant \mu(C) \leqslant \nu(C^\epsilon)+\epsilon \leqslant 
\nu(B^\epsilon)+\epsilon.
$$
Therefore $\mu(B) \leqslant \nu(B^\epsilon)+\epsilon$ since $\delta$ was arbitrary. 
The other inequality is similarly proved.
&square;


**Lemma.** *Let $E$ be a metric space, and $\mu\in\Pr(E)$. 
For every $\eta>0$, one can cover $E$ by open balls $\{S_i\}$ 
satisfying $\mu(\partial S_i)=0$ and $\textrm{diam}(S_i) < \eta$ for all $i$. 
If $E$ is separable, one can cover $E$ by countably many such balls.*

*Proof.* Denote by $S_\eta(x)$ the open sphere of radius $\eta$ centered 
at $x$, and by 
$B_\eta(x)=\partial S_\eta(x) = \bigl\{y \mid d(x,y)=\eta\bigr\}$ the 
boundary of this sphere. 
One has $S_\eta(x) = \bigcup_{0\leqslant\eta'<\eta}B_\eta(x)$. 
This is an uncountable disjoint union of Borel sets, and hence all but a 
countable number of them must have $\mu$-measure zero. 
Therefore every sphere $S_\eta(x)$ contains a sphere $S_{\eta'}(x)$, 
$\eta'\leqslant\eta$ such that $\mu\bigl(B_{\eta'}(x)\bigr)=0$. 
Picking such a sphere for every $x\in E$, one gets the first claim of the 
lemma. If $E$ is separable, it is Lindelöf, and the last statement follows.
&square;


**Theorem.** *The Prohorov distance defines a metric on $\Pr(E)$. 
The topology it induces is finer than the narrow topology. 
These two topologies coincide when $E$ is separable.*

*Proof.* 
It is clear that $d_P(\mu,\nu) = d_P(\nu,\mu) \geqslant 0$ and $d_P(\mu,\mu)=0$. 
If $A \subset E$ is closed then $A^\epsilon \downarrow A$ as 
$\epsilon \downarrow 0$ (it is easy to see that $d(x,A)=0$ for every 
$x \in \cap_{\epsilon>0} A^\epsilon$, and this is equivalent 
to $x\in\overline{A}$), hence $\mu(A^\epsilon) \downarrow \mu(A)$. 
Therefore $d_P(\mu,\nu) = 0$ implies $\mu(A)=\nu(A)$ for all closed $A$, 
and therefore $\mu = \nu$. Now let us prove the triangle inequality. 
Let $\mu_1,\mu_2,\mu_3\in\Pr(E)$. For any $\epsilon>d_P(\mu_1,\mu_2)$ and 
$\eta>d_P(\mu_2,\mu_3)$, we have
$$
\mu_1(A) < \mu_2(A^\epsilon)+\epsilon, \quad 
\mu_2(A^\epsilon) < \mu_3(A^{\epsilon+\eta})+\eta
$$
for every Borel set $A\subset E$. This implies 
$\mu_1(A) < \mu_3(A^{\epsilon+\eta})+\epsilon+\eta$. 
Similarly, $\mu_3(A) < \mu_1(A^{\epsilon+\eta})+\epsilon+\eta$. 
Therefore $d_P(\mu_1,\mu_3) \leqslant \epsilon+\delta$. 
Letting $\epsilon\to d_P(\mu_1,\mu_2)$ and $\eta\to d_P(\mu_2,\mu_3)$, we get
$d_P(\mu_1,\mu_3) \leqslant d_P(\mu_1,\mu_2) + d_P(\mu_2,\mu_3)$. 

Now let us show that the topology induced by $d_P$ is finer than 
the narrow topology. 

Let $\mu\in\Pr(E)$, $F\subset E$ closed and $\epsilon>0$. 
Let $\eta \in ]0,\epsilon[$ such that $\mu(F^\eta)<\mu(F)+\epsilon$. 
If $d_P(\mu,\nu)<\eta$, then 
$\nu(F) <\mu(F^\eta)+\eta < \mu(F) + 2\epsilon$. Thus each set of the family 
*(i)* in proposition *"neighborhood bases of the narrow topology"* contains a 
$d_P$-ball. The separability assumption has not been used here. 

Now we show the converse under the separability assumption on $E$ . 
We will show that every open $d_P$-ball contains a set of the family 
*(iii)* in proposition *"neighborhood bases of the narrow topology"*. 
Take $\eta \in ]0,\epsilon/2[$. Apply the previous lemma: 
cover $E$ by open balls ${\{S_i\}}_{i \geqslant 1}$ 
satisfying $\mu(\partial S_i)=0$ and $\textrm{diam}(S_i) < \eta$ for all $i$. 
Let $A_1 = S_1$ and 
$A_n = S_n\setminus\bigl(\bigcup_{m=1}^{n-1}S_m\bigr)$ for $n \geqslant 2$. 
Take $k\geqslant 1$ such that $\mu\bigl(\bigcup_{i=1}^kA_i\bigr) > 1-\eta$. 
One has $\bigcup_{i=1}^kA_i = \bigcup_{i=1}^kS_i$, therefore 
$\mu\Bigl(\partial\bigl(\bigcup_{i=1}^kA_i\bigr)\Bigr) = 0$.
Denote by $\mathcal{A}$ the set of all Borel $A$ that can be written as a 
union of sets from $\{A_1, \ldots, A_k\}$. 
Let $N$ be a neighborhood of $\mu$ of type *(iii)* in proposition 
*"neighborhood bases of the narrow topology"*, described by
$$
N = \Bigl\{\nu \in \Pr(E) \mid \forall A\in\mathcal{A}, 
\bigl|\mu(A)-\nu(A)\bigr|<\eta \Bigr\}.
$$
For any Borel set $B\subset E$, let $A'$ be the union of sets in 
$\{A_1, \ldots, A_k\}$ which intersect $B$. 
Then $\bigl|\mu(A')-\nu(A')\bigr|<\eta$ if $\nu \in N$.  
Let us check that $A'\subset B^\eta$. 
If $x \in A'$, then there is $j$ such that $x \in A_j$ and 
$A_j\cap B \neq \varnothing$. 
Pick $y \in A_j \cap B$. Then $d(x,B) \leqslant d(x,y)$. 
Since $\textrm{diam}(A_j)<\eta$, one has $x \in B^\eta$. 

One has $B \subset A' \cup {\bigl(\bigcup_{i=1}^kA_i\bigr)}^c$. Thus
$$
\mu(B) \leqslant \mu(A') + \mu\Bigl(\bigl(\bigcup_{i=1}^kA_i\bigr)^c\Bigr) 
\leqslant \nu(A') + 2\eta \leqslant \nu(B^\eta)+2\eta,
$$
hence $d_P(\mu,\nu) \leqslant 2\eta < \epsilon$. That is, every open $d_P$-ball 
contains a set of type *(iii)* in proposition 
*"neighborhood bases of the narrow topology"*.
&square;


We refer the reader to [3] for a proof of the following theorem. 
But we will give a proof of this theorem in the last section for a 
Lusin space $E$, which is enough for our purpose (since a Polish 
space is Lusin).

**Theorem (Prohorov).** *Let $E$ be a metrizable space. 
If $\Lambda \subset \Pr(E)$ is tight, then every sequence in $\Lambda$ has 
a convergent subsequence.*


**Theorem.** *If $E$ is Polish, the Prohorov metric is complete*.

*Proof.*
Let $(\mu_n)$ be a Cauchy sequence in $\Pr(E)$ with respect to $d_P$. 
Let $\epsilon>0$ and $k \geqslant 1$ be an integer. 
Pick $n(k)\geqslant 1$ such that 
$d_P(\mu_n,\mu_{n(k)}) < \epsilon 2^{-k}$ for $n \geqslant n(k)$. 
Consider a complete metric on $E$ compatible with the topology. 
Since $E$ is separable, it can be covered by countably many open balls 
of diameter $\epsilon 2^{-k}$. 
Let $B_1^k, \ldots, B_{m(k)}^k$ be finitely many open balls of diameter 
$\epsilon 2^{-k}$ satisfying
$$
\mu_{n(k)}\left(\bigcup_{i=1}^{m(k)}B_i^k\right) > 1 - \frac{\epsilon}{2^k}.
$$
For each $i \in [\![1,m(k)]\!]$, let $A_i^k$ be the open ball concentric with 
$B_i^k$ and with twice the radius. Since
$$
{\left(\bigcup_{i=1}^{m(k)}B_i^k\right)}^{\!\!\frac{\epsilon}{2^k}}
\subset \bigcup_{i=1}^{m(k)}A_i^k
$$
and $d_P(\mu_n,\mu_{n(k)}) < \frac{\epsilon}{2^k}$ for $n\geqslant n(k)$, 
$$
\mu_n\left(\bigcup_{i=1}^{m(k)}A_i^k\right) > 
1 - \frac{2\epsilon}{2^k}
$$
for $n\geqslant n(k)$. 

By adding if necessary finitely many open balls 
$A^k_{m(k)+1}, \ldots, A^k_{j(k)}$ of radius $\frac{2\epsilon}{2^k}$, we have
$$
\mu_n\left(\bigcup_{i=1}^{j(k)}A_i^k\right) > 
1 - \frac{2\epsilon}{2^k}
$$
for $n\geqslant 1$. 
It is easy to see that the set
$$
K = \bigcap_{k\geqslant 1}\bigcup_{i=1}^{j(k)}A_i^k
$$
is totally bounded. Therefore its closure $\overline{K}$ is totally bounded 
as well. Since it is complete, $\overline{K}$ is compact. 
Also, $\mu_n(\overline{K}) \geqslant 1-2\epsilon$ for all $n$. 
Therefore the sequence $(\mu_n)$ is tight. By Prohorov's theorem, it has 
a convergent subsequence. Since it is Cauchy, it converges.
&square;


**Corollary.** *If $E$ is Polish, $\Pr(E)$ is Polish.*

*Proof.* This is a consequence of the two previous theorems. 
&square;


# Kernels as random probabilities

Now we show the result stated in the introduction, namely that a 
probability kernel from a probability space to a Polish space $E$ is 
a random variable in the Polish space $\Pr(E)$. 

**Lemma.** *Let $E$ be a Polish space and $\Omega$ be a probability space. 
Then a map $\Gamma\colon\Omega\to\Pr(E)$ is measurable if and only if 
$\omega\mapsto\Gamma_\omega(f)$ is measurable for every $f\in C_b(E)$, 
where we denote by $\Gamma_\omega$ the probability 
$\mathcal{B}_E \ni A \mapsto \Gamma(\omega)(A)$.*

*Proof.* 
The "only if" part is obvious (and does not require the Polish assumption). 
Let us show the converse. Since $E$ is Polish, $\Pr(E)$ is Polish, hence it is 
strongly Lindelöf. 
Therefore every open set $O \subset \Pr(E)$ is a countable union of sets 
$$
\Bigl\{\nu \mid \bigl|\nu(f_i) - \mu(f_i)\bigr| < \epsilon, 
i\in [\![1,k]\!] \Bigr\}
$$
where $\mu\in\Pr(E)$, $\epsilon>0$, $k\geqslant 1$ and $f_i \in C_b(E)$. 
The maps $\omega\mapsto\Gamma_\omega(f_i)$ are measurable, therefore the set 
$$
\Bigl\{\omega \mid 
\bigl|\Gamma_\omega(f_i) - \mu(f_i)\bigr|<\epsilon, i\in [\![1,k]\!] \Bigr\}
$$
is measurable. Thus $\Gamma^{-1}(O)$ is a measurable set.
&square;


**Theorem.** *Let $E$ be a metric space. 
Let $\Omega$ be a probability space, and $\Gamma\colon\Omega\to\Pr(E)$ be a map. 
We denote by $\Gamma_\omega$ the image of $\omega$ by $\Gamma$. 
Define the map  $K\colon\Omega\times\mathcal{B}_E \to [0,1]$ by 
$K(\omega,A) = \Gamma_\omega(A)$. 
If $\Gamma$ is measurable, then $K$ is a probability kernel 
from $\Omega$ to $E$. 
If $E$ is Polish, the converse is true.* 

*Proof.*
Assume $\Gamma$ is measurable. Then $\omega\mapsto\Gamma_\omega(f)$ is 
measurable for every $f \in C_b(E)$. 
Let $F\subset E$ be a closed set and for every integer $n\geqslant 1$ 
define $f_n\in C_b(E)$ by $f_n(x)=\max\bigl\{0,1-n d(x,F)\bigr\}$ 
where $d$ is the metric on $E$. 
Then $f_n(x)\downarrow\mathbf{1}_F(x)$ for every $x\in E$. 
By monotone convergence, 
$\Gamma_\omega(f_n) \to \Gamma_\omega(F)$ and then 
$\omega\mapsto\Gamma_\omega(F)$ is measurable. It is easy to see that the set
$$
\bigl\{A \in \mathcal{B}_E \mid 
\omega\mapsto\Gamma_\omega(A) \text{ is measurable}\bigr\}
$$
is a $\lambda$-system. Since it contains the $\pi$-system of closed sets, 
it is equal to $\mathcal{B}_E$ by the monotone class theorem.

Now assume that $E$ is Polish and $K$ is a probability kernel. The set 
$$
V = \bigl\{f\colon E \to \mathbb{R} \mid 
\omega\mapsto\Gamma_\omega(f) \text{ is measurable}\bigr\}
$$
is a vector space containing the indicator function $\mathbf{1}_A$ for 
every $A\in\mathcal{B}_E$, hence it contains all finite linear combinations 
of such indicators functions. 
Let $f\in C_b(E)$. For every integer $n\geqslant 1$, define
$$
f_n(x) = \sum_{j=1}^\infty \frac{j-1}{n} 
\mathbf{1}_{\left]\frac{j-1}{n},\frac{j}{n}\right]}\bigl(f(x)\bigr) = 
\sum_{j=1}^\infty \frac{j-1}{n} 
\mathbf{1}_{f^{-1}\left(\left]\frac{j-1}{n},\frac{j}{n}\right]\right)}(x).
$$
Then $f_n$ is a finite linear combination since $f$ is bounded, and 
$\bigl|f_n(x)-f(x)\bigr| \leqslant \frac{1}{n}$ for every $x\in E$. 
Therefore $f_n(x) \leqslant 1+{\Vert f \Vert}_\infty$ and $f_n(x) \to f(x)$ 
for every $x\in E$. 
By dominated convergence, $\Gamma_\omega(f_n) \to \Gamma_\omega(f)$, 
hence $f \in V$. By the previous lemma, $\Gamma$ is measurable.
&square; 


# References 

- [1] Nikolaos S. Papageorgiou, Patrick Winkert. 
*Applied Nonlinear Functional Analysis: An Introduction*. 2018.

- [2] Mícheál Ó Searcóid. *Metric spaces.* 2007. 

- [3] Patrick Billingsley. 
*Convergence of probability measures.* Second edition, 1999.

- [4] K. R. Parthasarathy. 
*Probability Measures on Metric Spaces.* 1967.

- [5] Vivek S. Borkar. 
*Probability Theory: An Advanced Course.* 1995.

- [6] Alexander S. Kechris. 
*Classical descriptive set theory.* 1995.

