---
title: "Fast expansion of a polynomial with R"
author: "Stéphane Laurent"
date: '2023-04-29'
tags: R, maths
rbloggers: yes
output:
  md_document:
    variant: markdown
    preserve_yaml: true
  html_document:
    highlight: kate
    keep_md: no
highlighter: pandoc-solarized
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, message = FALSE)
```


In [this previous post](https://laustep.github.io/stlahblog/posts/caracas01.html), 
[this previous post](https://laustep.github.io/stlahblog/posts/caracas02.html), 
and [this previous post](https://laustep.github.io/stlahblog/posts/juliaPolynomialExpansion.html), 
I showed how to expand a polynomial with symbolic parameters. 
In the first two posts, I used the R package **caracas**, a wrapper of 
**SymPy**, and in the third post I used Julia.

Now I've found a pure R solution, with the help of the **spray** package, and 
it is very fast. That's what I'm going to demonstrate here, with the same 
example I treated with Julia.

We will lost something as compared to Julia: the **spray** package does not 
handle rational numbers, so we will replace them by their double approximation. 
We could use the **qspray** package instead, to deal with the rational numbers.

Here is the starting polynomial expression:

```{r}
# define the polynomial expression as a function
f <- function(x, y, z, w, sqrt3) {
  ((x^2+y^2+z^2+w^2+145/3)^2-4*(9*z^2+16*w^2))^2 * 
    ((x^2+y^2+z^2+w^2+145/3)^2+296*(x^2+y^2)-4*(9*z^2+16*w^2)) - 
    16*(x^2+y^2)*(x^2+y^2+z^2+w^2+145/3)^2 * 
    (37*(x^2+y^2+z^2+w^2+145/3)^2-1369*(x^2+y^2)-7*(225*z^2+448*w^2)) - 
    sqrt3*16/9*(x^3-3*x*y^2) * 
    (110*(x^2+y^2+z^2+w^2+145/3)^3-148*(x^2+y^2+z^2+w^2+145/3) * 
       (110*x^2+110*y^2-297*z^2+480*w^2)) - 
    64*(x^2+y^2)*(3*(729*z^4+4096*w^4)+168*(x^2+y^2)*(15*z^2-22*w^2)) + 
    64*(12100/27*(x^3-3*x*y^2)^2-7056*(3*x^2*y-y^3)^2)-592240896*z^2*w^2
}
```

We transform it to the polynomial of interest:

```{r}
# there are 3 variables (x,y,z) and 6 parameters (w0,a,b,c,d,sqrt3)
library(spray)
x     <- lone(1, 9)
y     <- lone(2, 9)
z     <- lone(3, 9)
w0    <- lone(4, 9)
a     <- lone(5, 9)
b     <- lone(6, 9)
c     <- lone(7, 9)
d     <- lone(8, 9)
sqrt3 <- lone(9, 9)

# define the substitutions
X <- a*x  - b*y  - c*z  - d*w0
Y <- a*y  + b*x  + c*w0 - d*z
Z <- a*z  - b*w0 + c*x  + d*y
W <- a*w0 + b*z  - c*y  + d*x

# here is the polynomial of interest
P <- f(X, Y, Z, W, sqrt3)
```

Now we take the "xyz" part of each term:

```{r}
# the exponents of x, y, z
xyz_powers <- P[["index"]][, c(1L, 2L, 3L)]

# the "keys" xyz(i,j,k) for POV-Ray
xyz_povray <- apply(xyz_powers, 1L, function(comp) {
  sprintf("xyz(%s): ", toString(comp))
})
```

Now we take the remaining part of each term:

```{r}
# the other exponents, those of w0, a, b, c, d, sqrt3
other_powers <- P[["index"]][, c(4L, 5L, 6L, 7L, 8L, 9L)]

# the polynomials in w0, a, b, c, d, sqrt3
nterms <- length(P)
coeffs <- P[["value"]]
polynomials <- lapply(1L:nterms, function(i) {
  spray(other_powers[i, ], coeffs[i])
})
```

And we group these polynomials, putting together those which share the 
same "xyz" part:

```{r}
# group the polynomials which have the same x,y,z exponents
polynomials_groups <- sapply(split(polynomials, xyz_povray), function(polys) {
  polysum <- polys[[1]]
  for(poly in polys[-1]) {
    polysum <- spray_add(
      polysum$index, polysum$value, poly$index, poly$value
    )
  }
  as.spray(polysum)
}, simplify = FALSE)
# remove the empty polynomials, if there are some
polynomials_groups <- Filter(Negate(is.empty), polynomials_groups)
```

We want these polynomials as character strings:

```{r}
# there's no as.character function for sprays (polynomials), so we make one
asCharacter <- function(poly) {
  op <- options(sprayvars = c("w0", "a", "b", "c", "d", "sqrt3"))
  x <- capture.output(print_spray_polyform(poly))
  options(op)
  paste0(x, collapse = " ")
}

polynomials_strings <- 
  sapply(polynomials_groups, asCharacter, simplify = FALSE)
```

And that's it. We have everything needed to construct the POV-Ray code:

```{r}
head(polynomials_strings, 2L)
```

