---
title: "Calling a Haskell function in R - a float expansion example"
author: "Stéphane Laurent"
date: '2017-06-03'
output:
  md_document:
    toc: yes
    variant: markdown
  html_document:
    highlight: zenburn
    keep_md: yes
prettify: yes
prettifycss: minimal
highlighter: pandoc-solarized
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse=TRUE)
```


In [the previous article](https://laustep.github.io/stlahblog/posts/DyadicExpansion.html), I wrote a R function returning the binary expansion of a real number in $[0,1]$. In the present article, I will: 

- write a similar function in Haskell;
- write this function in a way compatible with R, inside a module;
- compile this module in a dynamic linker suitable for R (`dll` for Windows, `so` for Linux);
- call the function from R through the dynamic linker.

The creation of a Haskell function compatible with R is allowed by the [Foreign Function Interface (FFI)](https://wiki.haskell.org/Foreign_Function_Interface), in other words the `Foreign` module.

I learnt how to do such things with the help of [this blog post by Neil Mitchell](http://neilmitchell.blogspot.be/2011/10/calling-haskell-from-r.html).


## Binary (and more) expansion in Haskell

Let's go to Haskell.
The `floatExpansion` function below is obtained by a small modification of the `floatToDigits` function of the `Numeric` module. It returns the expansion of a real number $u \in [0,1]$ in a given integer base.

```{r, engine='haskell', engine.path='ghc'}
import Numeric (floatToDigits)
:{
let floatExpansion :: RealFloat a => Integer -> a -> [Int];
    floatExpansion base u = replicate (- snd expansion) 0 ++ fst expansion
      where expansion = floatToDigits base u
:}
floatExpansion 2 0.125
```


## First dynamic linker: string output

Firstly, I show how to make this function compatible with R when its output is a string instead of a list. It is easy to convert a list to a string in Haskell:

```{r, engine='haskell', engine.path='ghc'}
show [0, 0, 1]
```

To get the output as a vector in R, more work is needed, and I will do it in the next section.


### Make the function compatible with R

To make the function compatible with R, there are two rules:

- Every argument must be a pointer (`Ptr`) to a C compatible type: `CInt`, `CDouble` or `CString`.

- The result must be `IO ()`.

A value of type `Ptr` represents a pointer to an object. This type is provided by the [`Foreign.Ptr` module](https://hackage.haskell.org/package/base-4.9.0.0/docs/Foreign-Ptr.html), which is imported via the `Foreign` module.
The types `CInt`, `CDouble` and `CString` are provided by the [`Foreign.C` module](https://hackage.haskell.org/package/base-4.9.0.0/docs/Foreign-C.html).

We end up with this module:

```haskell
-- FloatExpansion1.hs
{-# LANGUAGE ForeignFunctionInterface #-}

module FloatExpansion where

import Foreign
import Foreign.C
import Numeric (floatToDigits)

foreign export ccall floatExpansion :: Ptr CInt -> Ptr CDouble -> Ptr CString -> IO ()

floatExpansion :: Ptr CInt -> Ptr CDouble -> Ptr CString -> IO ()
floatExpansion base u result = do
  base <- peek base
  u <- peek u
  expansion <- newCString $ show $ floatExpansion' (toInteger base) u
  poke result $ expansion

floatExpansion' :: RealFloat a => Integer -> a -> [Int]
floatExpansion' base u = replicate (- snd expansion) 0 ++ fst expansion
  where expansion = floatToDigits base u
```


### Compilation

We need the following C file to do the compilation, as explained in the [GHC users guide](https://downloads.haskell.org/~ghc/latest/docs/html/users_guide/win32-dlls.html#making-dlls-to-be-called-from-other-languages).

```c
// StartEnd.c
#include <Rts.h>

void HsStart()
{
int argc = 1;
char* argv[] = {"ghcDll", NULL}; // argv must end with NULL

// Initialize Haskell runtime
char** args = argv;
hs_init(&argc, &args);
}

void HsEnd()
{
hs_exit();
}
```

Then we compile the library with this command on Linux:

```bash
ghc -shared -fPIC -dynamic -lHSrts-ghc8.0.2 FloatExpansion1.hs StartEnd.c -o FloatExpansion1.so
```

and this command on Windows:

```bash
ghc -shared -fPIC FloatExpansion1.hs StartEnd.c -o FloatExpansion1.dll
```

This creates the dynamic linker `FloatExpansion1.so` on Linux, `FloatExpansion1.dll` on Windows.

In a cabal file, assuming `StartEnd.c` is in the project directory, we can do:

```cabal
library
  hs-source-dirs:      src
  exposed-modules:     FloatExpansion
  build-depends:       base >= 4.7 && < 5
  default-language:    Haskell2010
  if os(windows)
    ghc-options:       -O2 -shared -fPIC StartEnd.c -o FloatExpansion1.dll
  else
    ghc-options:       -O2 -shared -fPIC -dynamic StartEnd.c -o FloatExpansion1.so
    extra-libraries:   HSrts-ghc8.0.2
```


### Call in R

We firstly load the library with:

```{r, cache=TRUE}
dyn.load("FloatExpansion1.dll")
.C("HsStart")
```

And we invoke the function with the help of the `.C` function, as follows:

```{r, cache=TRUE}
.C("floatExpansion", base=2L, x=0.125, result="")$result
```

It works. But it would be better to have a vector as output, rather than a string.

```{r, cache=TRUE}
dyn.unload("FloatExpansion1.dll")
```


## Second dynamic linker: vector output

To get the output as a vector, the additional modules we need are: `Foreign.R`, `Foreign.R.Types` and `Data.Vector.SEXP`. They are provided by the [`inline-r` package](https://hackage.haskell.org/package/inline-r).
The `[Int]` type of the output list of the `floatExpansion` function must be converted to `[Int32]`. We write a simple function `intToInt32` to help us to do the conversion. It works with the help of the `Data.Int` module which is imported via the `Foreign` module.

We end up with this module:

```haskell
-- FloatExpansion2.hs
{-# LANGUAGE ForeignFunctionInterface #-}
{-# LANGUAGE DataKinds #-}

module FloatExpansion where

import Foreign
import Foreign.C
import Foreign.R (SEXP)
import qualified Foreign.R.Type as R
import qualified Data.Vector.SEXP as DV
import Numeric (floatToDigits)

foreign export ccall floatExpansion :: Ptr CInt -> Ptr CDouble -> Ptr (SEXP s R.Int) -> IO ()

floatExpansion :: Ptr CInt -> Ptr CDouble -> Ptr (SEXP s R.Int) -> IO ()
floatExpansion base u result = do
  base <- peek base
  u <- peek u
  let expansion = map intToInt32 $ floatExpansion' (toInteger base) u
  poke result $ DV.toSEXP $ DV.fromList expansion

intToInt32 :: Int -> Int32
intToInt32 = fromIntegral

floatExpansion' :: RealFloat a => Integer -> a -> [Int]
floatExpansion' base u = replicate (- snd expansion) 0 ++ fst expansion
  where expansion = floatToDigits base u
```

We compile the library as before. And we load it in R as before:

```{r}
dyn.load("FloatExpansion2.dll")
.C("HsStart")
```

And we invoke the function with the help of the `.C` function, as follows:

```{r}
.C("floatExpansion", base=2L, x=0.125, result=list(0L))$result
```

In fact, the output is a list with one element, the desired vector.

Let's write a user-friendly function:

```{r}
floatExpand <- function(x, base=2L){
  .C("floatExpansion", base=as.integer(base), x=as.double(x), result=list(integer(1)))$result[[1]]  
}
```

```{r, include=FALSE}
num2dyadic <- function(u, nmax=1024L){
  out <- integer(nmax)
  i <- j <- 0L
  while(u>0 && i<nmax){
    j <- 1L + max(0L,floor(-log2(u*(1+.Machine$double.eps^.5))))
    if(i+j <= nmax){
      i <- i + j
      out[i] <- 1L
      u <- 2L^j*u - 1L
    }else{
      i <- nmax
    }
  }
  return(out[1:i])
}
```

Let's compare it with my R function `num2dyadic`:

```{r}
library(microbenchmark)
microbenchmark(
  floatExpand = floatExpand(runif(1)),
  num2dyadic  = num2dyadic(runif(1)),
  times = 5000
)
```

It is faster. And I have checked that the two functions always return the same results.

Moreover the "RHaskell" function allows more than the binary expansion, for example the ternary expansion:

```{r}
floatExpand(1/3+1/27, base=3)
```

Quite nice, isn't it ?

```{r, eval=FALSE}
dyn.unload("FloatExpansion2.dll")
```
