---
title: "Generalized circles and inversions"
author: "Stéphane Laurent"
date: '2019-03-13'
tags: R, maths, geometry
output:
  md_document:
    variant: markdown
    preserve_yaml: true
  html_document:
    highlight: kate
    keep_md: no
prettify: yes
linenums: yes
prettifycss: minimal
highlighter: pandoc-solarized
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, fig.path="./figures/Inversions-")
```

Let $\iota_0$ be the inversion with respect to the unit circle:
$$
\iota_0(z) = \frac{1}{\bar{z}}, \quad z \neq 0.
$$

Consider a 
[generalized circle](https://laustep.github.io/stlahblog/posts/MobiusTransformations.html) 
$\mathcal{G}\mathcal{C}$ having equation
$$
\boxed{Az\bar{z} + \bar{\gamma}z + \gamma\bar{z} = D}
$$
which is equivalent, by dividing by $z\bar{z}$, to
$$
A + \bar{\gamma}\frac{1}{\bar{z}} + \gamma\frac{1}{z} = \frac{D}{z\bar{z}}.
$$

Plugging $z = \iota_0^{-1}(w) = 1/\bar{w}$ into this equation, we get the 
following equation of $\iota_0(\mathcal{G}\mathcal{C})$:
$$
A + \bar{\gamma}w + \gamma\bar{w} = D w\bar{w} \\
\iff \, 
\boxed{D w\bar{w} - \bar{\gamma}w - \gamma\bar{w} = A}.
$$
This is the equation of a generalized circle.

Now consider the inversion $\iota$ with respect to the circle of center $c$ 
and radius $r$:
$$
\iota(z) = c + \frac{r^2}{\overline{z-c}}, \quad z \neq c.
$$

One has
$$
\iota = f^{-1} \circ \iota_0 \circ f
$$
where $f(z) = \frac{z-c}{r}$, and $f^{-1}(z) = rz + c$.

The transformation $f$ is a "scaling-translation". It is clear that it maps a 
generalized circle to a generalized circle, and this also holds for $f^{-1}$. 
Therefore, $\iota$ maps any generalized circle to a generalized circle. 
This result is well-known. 
Here, we will give the equation of the image generalized circle.


# Case when $\mathcal{G}\mathcal{C}$ is a circle 

Assume that $\mathcal{G}\mathcal{C}$ is the circle $\mathcal{C}_0$ with center 
$z_0$ and radius $R_0$. 
Then $f(\mathcal{C_0})$ is the circle $\mathcal{C_1}$ with center 
$z_1 = \frac{z_0-c}{r}$ and radius $R_1 = \frac{R_0}{r}$. It has equation
$$
z\bar{z} - \bar{z_1}z - z_1\bar{z} = D_1
$$
with $D_1 = R_1^2-{|z_1|}^2$.

Then $\iota_0(\mathcal{C_1})$ is the generalized circle having equation
$$
D_1 z\bar{z} + \bar{z_1}z + z_1\bar{z} = 1.
$$


### Case $D_1 \neq 0$

If $D_1 \neq 0$, the above equation of $\iota_0(\mathcal{C_1})$ is the 
equation of a circle $\mathcal{C}_2$ with center $z_2 = -\frac{z_1}{D_1}$ 
and radius 
$$
R_2 = \sqrt{\frac{{|z_1|}^2}{D_1^2} + \frac{1}{D_1}}.
$$

Now, $f^{-1}(\mathcal{C}_2) = \iota(\mathcal{C}_0)$ is the circle with center 
$r z_2 + c$ and radius $r R_2$. 


### Case $D_1 = 0$

If $D_1=0$, our equation of $\iota_0(\mathcal{C_1})$ is 
$$
\bar{z_1}z + z_1\bar{z} = 1,
$$
which is the equation of a line $\mathcal{L_2}$, with direction 
$\theta_2 = \textrm{arg}(z_1)$ and offset $d_2 = \frac{1}{2{|z_1|}^2}$.

Then $f^{-1}(\mathcal{L}_2) = \iota(\mathcal{C}_0)$ is a line parallel to 
$\mathcal{L}_2$, that is, its direction is $\theta_2$. 
To get its offset, take a point $w_2$ on $\mathcal{L}_2$:
$$
w_2 = \begin{cases}
\dfrac{1}{2 \Re(z_1)} & \text{if } \Im(z_1) = 0 \\
\dfrac{i}{2 \Im(z_1)} & \text{if } \Im(z_1) \neq 0
\end{cases}
$$
and take its image $w_3 = f^{-1}(w_2) = r w_2 + c$. Then the offset is 
$$
\cos(\theta_2) \Re(w_3) + \sin(\theta_2) \Im(w_3) \\
\qquad = \frac{\Re(z_1)}{|z_1|}\Re(w_3) +  \frac{\Im(z_1)}{|z_1|}\Im(w_3).
$$


# Case when $\mathcal{G}\mathcal{C}$ is a line

Now assume that $\mathcal{G}\mathcal{C}$ is the line $\mathcal{L}_0$ with 
direction $\theta_0$ and offset $d_0$. 
Its image $f(\mathcal{L}_0)$ is a line $\mathcal{L}_1$ parallel to 
$\mathcal{L}_0$, that is the direction of $\mathcal{L}_1$ is $\theta_0$. 
To get its offset, take a point $w_0$ on $\mathcal{L}_0$:
$$
w_0 = \begin{cases}
\dfrac{d_0}{\cos(\theta_0)} & \text{if } \sin(\theta_0) = 0 \\
i\dfrac{d_0}{\sin(\theta_0)} & \text{if } \sin(\theta_0) \neq 0
\end{cases}
$$
and take its image $w_1 = f(w_0) = \frac{w_0-c}{r}$. Then the offset of 
$\mathcal{L}_1$ is 
$$
d_1 = \cos(\theta_0) \Re(w_1) + \sin(\theta_0) \Im(w_1)
$$
and the line $\mathcal{L}_1$ has equation
$$
\bar{\gamma_1}z + \gamma_1\bar{z} = D_1 
$$
with $\gamma_1 = \cos(\theta_0) + i\sin(\theta_0)$ and $D_1 = 2d_1$.

Therefore, $\iota_0(\mathcal{L}_1)$ is the generalized circle having equation
$$
D_1 z\bar{z} - \bar{\gamma_1}z - \gamma_1\bar{z} = 0.
$$


### Case $D_1 \neq 0$

If $D_1 \neq 0$, then $\iota_0(\mathcal{L}_1)$ is the circle $\mathcal{C}_2$ 
with center $z_2 = \frac{\gamma_1}{D_1}$ and radius 
$R_2 = \frac{|\gamma_1|}{|D_1|}$. 
Then $f^{-1}(\mathcal{C}_2) = \iota(\mathcal{L}_0)$ is the circle with 
center $r z_2 + c$ and radius $r R_2$.


### Case $D_1 = 0$

If $D_1 = 0$, then $\iota_0(\mathcal{L}_1)$ is the line $\mathcal{L}_2$ having 
equation 
$$
\bar{\gamma_1}z + \gamma_1\bar{z} = 0.
$$
Its direction is $\textrm{arg}(\gamma_1) = \theta_0$ and its offset is $0$.

Then $f^{-1}(\mathcal{L}_2) = \iota(\mathcal{C}_0)$ is a line $\mathcal{L}_3$ 
parallel to $\mathcal{L}_2$, that is, its direction is $\theta_0$. 
The line $\mathcal{L}_2$ passes through $0$, therefore the point 
$f^{-1}(0) = c$ belongs to $\mathcal{L}_3$. 
Consequently, the offset of $\mathcal{L}_3$ is 
$$
\cos(\theta_0) \Re(c) + \sin(\theta_0) \Im(c).
$$


# R code 

We firstly write a function `iotaCircle` which calculates the image of 
a circle.

```{r iotaCircle}
# squared modulus
Mod2 <- function(z){
  Re(z)^2 + Im(z)^2
}
# image of a circle
iotaCircle <- function(c, r, circle, tol = sqrt(.Machine$double.eps)){
  z0 <- circle$center; R0 <- circle$radius
  z1 <- (z0-c)/r
  D1 <- (R0/r)^2 - Mod2(z1)
  if(abs(D1) > tol){
    z2 <- -z1/D1
    R2 <- sqrt(Mod2(z2) + 1/D1)
    out <- list(center = r*z2+c, radius = r*R2)
    attr(out, "type") <- "circle"
  }else{
    theta2 <- Arg(z1) %% (2*pi)
    w2 <- ifelse(Im(z1) == 0, 1/2/Re(z1), 1i/2/Im(z1))
    w3 <- r*w2 + c
    out <- list(
      theta = theta2, 
      offset = (Re(z1)*Re(w3)+Im(z1)*Im(w3))/Mod(z1)
    )
    attr(out, "type") <- "line"
  }
  out
}
```

Let's check the function `iotaCircle`. 
To do so, we use the `circumcircle` function below, which returns the center 
and the radius of the circle passing by `z1`, `z2`, `z3`.

```{r circumcircle}
circumcircle <- function(z1,z2,z3){
  x1 <- Re(z1); y1 <- Im(z1); p1 <- c(x1,y1)
  x2 <- Re(z2); y2 <- Im(z2); p2 <- c(x2,y2)
  x3 <- Re(z3); y3 <- Im(z3); p3 <- c(x3,y3)
  a <- det(cbind(rbind(p1,p2,p3),1))
  q1 <- c(crossprod(p1)); q2 <- c(crossprod(p2)); q3 <- c(crossprod(p3))
  q <- c(q1,q2,q3)
  x <- c(x1,x2,x3)
  y <- c(y1,y2,y3)
  Dx <- det(cbind(q,y,1))
  Dy <- -det(cbind(q,x,1))
  c <- det(cbind(q,x,y))
  center <- c(Dx,Dy)/a/2
  r <- sqrt(c(crossprod(center-p1)))
  list(center = center[1] + 1i*center[2], radius = r)
}
```

Now let's check `iotaCircle`.

```{r check_iotaCircle}
# define iota
c <- 1 + 2i; r <- 3
iota <- function(z){
  c + r^2/Conj(z-c)
}
# define C0 - a case when iota(C0) is a circle ####
z0 <- 2 + 3i; R0 <- 2
iotaCircle(c, r, circle = list(center = z0, radius = R0))
circumcircle(iota(z0+R0), iota(z0-R0), iota(z0+1i*R0))
# now, a case when iota(C0) is a line ####
R0 <- sqrt(2)
w1 <- iota(z0+R0); w2 <- iota(z0-R0); w3 <- iota(z0+1i*R0)
(w2-w1)/(w3-w1) # aligned <=> Im = 0
(L <- iotaCircle(c, r, circle = list(center = z0, radius = R0)))
cos(L$theta)*Re(w1) + sin(L$theta)*Im(w1) # = L$offset
cos(L$theta)*Re(w2) + sin(L$theta)*Im(w2) # = L$offset
cos(L$theta)*Re(w3) + sin(L$theta)*Im(w3) # = L$offset
```

Now we write a function `iotaLine` which calculates the image of a line.

```{r iotaLine}
iotaLine <- function(c, r, line, tol = sqrt(.Machine$double.eps)){
  theta0 <- line$theta; d0 <- line$offset
  w0 <- ifelse(sin(theta0) == 0, d0/cos(theta0), 1i*d0/sin(theta0))
  w1 <- (w0-c)/r
  D1 <- 2 * (cos(theta0)*Re(w1) + sin(theta0)*Im(w1))
  if(abs(D1)>tol){
    gamma1 <- cos(theta0) + 1i*sin(theta0)
    z2 <- gamma1/D1
    R2 <- Mod(gamma1/D1)
    out <- list(center = r*z2+c, radius = r*R2)
    attr(out, "type") <- "circle"
  }else{
    out <- list(
      theta = theta0, 
      offset = cos(theta0)*Re(c)+sin(theta0)*Im(c)
    )
    attr(out, "type") <- "line"
  }
  out
}
```

Let's check `iotaLine`. Firstly, a case when the image is a circle.

```{r check_iotaLine1}
# define L0
theta0 <- 1; d0 <- 2
theta0 <- pi/4; d0 <- 0
# take three points on iota(L0)
x <- c(0,1,2)
y <- (d0 - cos(theta0)*x) / sin(theta0)
w1 <- iota(x[1] + 1i*y[1])
w2 <- iota(x[2] + 1i*y[2])
w3 <- iota(x[3] + 1i*y[3])
# check
circumcircle(w1,w2,w3)
iotaLine(c, r, list(theta = theta0, offset = d0))
```

Now a case when the image is a line.

```{r check_iotaLine2}
# define iota
c <- 1-1i 
iota <- function(z){
  c + r^2/Conj(z-c)
}
# define L0
theta0 <- pi/4; d0 <- 0
# take three points on iota(L0)
x <- c(0,2,4)
y <- (d0 - cos(theta0)*x) / sin(theta0)
w1 <- iota(x[1] + 1i*y[1])
w2 <- iota(x[2] + 1i*y[2])
w3 <- iota(x[3] + 1i*y[3])
# check
(w2-w1)/(w3-w1) # aligned <=> Im = 0
( L <- iotaLine(c, r, list(theta = theta0, offset = d0)) )
cos(L$theta)*Re(w1) + sin(L$theta)*Im(w1) # = L$offset
cos(L$theta)*Re(w2) + sin(L$theta)*Im(w2) # = L$offset
cos(L$theta)*Re(w3) + sin(L$theta)*Im(w3) # = L$offset
```

Finally, let's put things together in a single function.

```{r iotaGcircle}
iotaGcircle <- function(c, r, gcircle){
  if(attr(gcircle, "type") == "circle"){
    iotaCircle(c, r, gcircle)
  }else{
    iotaLine(c, r, gcircle)
  }
}
```

# Illustration 

We will make a beautiful picture of inverted circles. 
This is borrowed from this 
[circle inversion gallery](http://xahlee.info/SpecialPlaneCurves_dir/InversionGallery_dir/inversionGallery.html).

Firstly, let's write a function which draws a generalized circle. 
We use the `draw.circle` function of the `plotrix` package.

```{r}
library(plotrix)
drawLine <- function(line, ...){
  theta <- line$theta; offset <- line$offset
  if(sin(theta) != 0){
    abline(a = offset/sin(theta), b = -1/tan(theta), ...)
  }else{
    abline(v = offset/cos(theta), ...)
  }
}
drawCircle <- function(circle, ...){
  draw.circle(Re(circle$center), Im(circle$center), circle$radius, ...)
}
drawGcircle <- function(gcircle, color = "black", ...){
  if(attr(gcircle, "type") == "circle"){
    drawCircle(gcircle, border = color, ...)
  }else{
    drawLine(gcircle, col = color, ...)
  }
}
```

We start with five circles.

```{r gen0}
# generation 0
gen0 <- sapply(c(0, pi/2, pi, 3*pi/2), function(beta){
  out <- list(center = cos(beta)+1i*sin(beta), radius = 1, gen = 0)
  attr(out, "type") <- "circle"
  out
}, simplify = FALSE)
gen0[[5]] <- list(center = 0+0i, radius = 2, gen = 0)
attr(gen0[[5]], "type") <- "circle"
# plot
par(bg = "black", mar = c(0,0,0,0))
plot(0, 0, type="n", xlim=c(-2.3,2.3), ylim=c(-2.3,2.3),   
     asp=1, axes=FALSE, xlab=NA, ylab=NA)
invisible(lapply(gen0, drawGcircle, color = "yellow", lwd = 2))
```

Now, we invert each of these five circles with respect to the other ones:

```{r gen1}
# generation 1
n0 <- length(gen0)
n1 <- n0*(n0-1)
gen1 <- vector("list", n1)
k <- 0
while(k < n1){
  for(j in 1:n0){
    for(i in 1:n0){
      if(i != j){
        k <- k+1
        gen1[[k]] <- 
          iotaGcircle(gen0[[i]]$center, gen0[[i]]$radius, gen0[[j]])
        gen1[[k]]$base <- i
        gen1[[k]]$gen <- 1
      }
    }
  }
}
```

We continue so on: we invert the obtained generalized circles with respect to 
the starting circles.

```{r gen23}
# generation 2
n2 <- n0*n1-n1
gen2 <- vector("list", n2)
k <- 0
while(k < n2){
  for(j in 1:n1){
    for(i in 1:n0){
      if(gen1[[j]]$base != i){
        k <- k+1
        gen2[[k]] <- 
          iotaGcircle(gen0[[i]]$center, gen0[[i]]$radius, gen1[[j]])
        gen2[[k]]$base <- i
        gen2[[k]]$gen <- 2
      }
    }
  }
}
# generation 3
n3 <- n0*n2-n2
gen3 <- vector("list", n3)
k <- 0
while(k < n3){
  for(j in 1:n2){
    for(i in 1:n0){
      if(gen2[[j]]$base != i){
        k <- k+1
        gen3[[k]] <- 
          iotaGcircle(gen0[[i]]$center, gen0[[i]]$radius, gen2[[j]])
        gen3[[k]]$gen <- 3
      }
    }
  }
}
```

Let's put all the obtained generalized circles together:

```{r}
gcircles <- c(gen0, gen1, gen2, gen3)
length(gcircles)
```

There are 425 generalized circles, but some of them are duplicated. 
To remove the duplicates, we use the `uniqueWith` function below. 
This function takes as arguments an atomic vector or a list `v` and a 
function `f` of two arguments, two elements of `v`, and which returns `TRUE` or 
`FALSE`, according to whether the two elements are considered as duplicates.

```{r uniqueWith}
uniqueWith <- function(v, f){
  size <- length(v)
  for(i in seq_len(size-1L)){
    j <- i + 1L
    while(j <= size){
      if(f(v[[j]],v[[i]])){
        v <- v[-j]
        size <- size - 1L
      }else{
        j <- j + 1L
      }
    }
  }
  v[1L:size]
}
# examples ####
v <- c(1,1,2,1,3,4,3,5,5)
uniqueWith(v, `==`)
uniqueWith(v, function(x,y) (x-y) %% 3 == 0)
v <- list(a="you", b="are", c="great")
uniqueWith(v, function(x,y) nchar(x) == nchar(y))
```

So let's define the function `f` which identifies two equal generalized 
circles. 

```{r f}
f <- function(gcircle1, gcircle2){
  if(attr(gcircle1, "type") == attr(gcircle2, "type")){
    if(attr(gcircle1, "type") == "circle"){
      Mod(gcircle1$center-gcircle2$center) < 1e-5 && 
        abs(gcircle1$radius-gcircle2$radius) < 1e-5
    }else{
      abs(cos(gcircle1$theta)-cos(gcircle2$theta)) < 1e-5 &&
        abs(sin(gcircle1$theta)-sin(gcircle2$theta)) < 1e-5 &&
        abs(gcircle1$offset-gcircle2$offset) < 1e-5
    }
  }else{
    FALSE
  }
}
```

And now, let's remove the duplicates:

```{r}
gcircles <- uniqueWith(gcircles, f)
length(gcircles)
```

Now we plot the generalized circles, with a color indicating the generation.

```{r picture}
draw <- function(gcircle, colors=rainbow(4), ...){
  drawGcircle(gcircle, color = colors[1+gcircle$gen], ...)
}
par(mar = c(0,0,0,0), bg = "black")
plot(0, 0, type="n", xlim=c(-2.3,2.3), ylim=c(-2.3,2.3),  
     asp=1, axes=FALSE, xlab=NA, ylab=NA)
invisible(lapply(gcircles, draw, lwd=2))
```

