---
title: "Inversion swapping two circles"
author: "Stéphane Laurent"
date: '2020-02-09'
tags: R, maths, geometry
output:
  md_document:
    variant: markdown
    preserve_yaml: true
  html_document:
    highlight: kate
    keep_md: no
prettify: yes
linenums: yes
prettifycss: minimal
highlighter: pandoc-solarized
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, 
                      fig.path = "figures/InversionSwappingCircles-")
```

Consider two circles $\mathcal{C}_1(O_1,r_1)$ and $\mathcal{C}_2(O_2,r_2)$. 
There is always an inversion which swaps $\mathcal{C}_1$ and $\mathcal{C}_2$ 
(which maps $\mathcal{C}_1$ to $\mathcal{C}_2$ and $\mathcal{C}_2$ to 
$\mathcal{C}_1$). 
When this is an inversion on a circle, this circle of inversion is called 
a *mid-circle* of $\mathcal{C}_1$ and $\mathcal{C}_2$. 

Hereafter, we set $\rho = \frac{r_1}{r_2}$.

# Congruent case

When $r_1 = r_2$, the two circles $\mathcal{C}_1$ and $\mathcal{C}_2$ can be 
swapped by an inversion on a line (that is, a reflection with respect to this 
line); obviously, this line is the perpendicular bisector of the segment 
joining the two centers $O_1$ and $O_2$. 
We could say that this line is a mid-circle, if 
we allow the mid-circle to be a generalized circle. 
If the two circles are tangent, this is the unique mid-circle.

When $r_1 = r_2$, the two circles $\mathcal{C}_1$ and $\mathcal{C}_2$ can be 
swapped by an inversion on a circle if they intersect each other at two 
points. The pole of this inversion is 
$I = \frac{1}{2}(O_1+O_2)$, and its power is $k = |IO_2^2 - r_2^2|$. 
That is, the circle of center $I$ and radius $\sqrt{k}$ is a mid-circle. 

When $r_1 = r_2$ and the two circles $\mathcal{C}_1$ and $\mathcal{C}_2$ do 
not intersect each other, there is no inversion on a circle which swaps these 
circles. But they can be swapped by an inversion with a negative power; 
the pole of this inversion is 
$I = \frac{1}{2}(O_1+O_2)$, and its power is $k = -|IO_2^2 - r_2^2|$. 

Let's check this third claim. In a [previous post](./Inversions.html), I 
gave a formula to calculate the image of a circle by an inversion on a circle. 
A more general formula for the image circle $\mathcal{C}'(O',r')$ of the 
circle $\mathcal{C}(O,r)$ by the inversion of pole $I$ and possibly negative 
power $k$, is: $O' = I + \frac{k}{d}(O-I)$ and 
$r' = \left|\frac{k}{d}\right|r$ at condition that 
$d := OI^2 - r^2 \neq 0$ (if $d=0$, the image of $\mathcal{C}$ is a line).

```{r}
invertCircle <- function(I, k, circle){
  O <- circle$center; r <- circle$radius
  d <- c(crossprod(O-I)) - r^2
  list(
    center = I + k/d*(O-I),
    radius = abs(k/d)*r
  )
}
# two congruent circles which do not intersect each other
C1 <- list(center = c(1,2), radius = 3)
C2 <- list(center = c(14,15), radius = 3)
# inversion which swaps the two circles
I <- (C1$center + C2$center) / 2
k <- -abs(c(crossprod(I-C2$center) - C2$radius^2))
# check
invertCircle(I, k, C1) # should be C2
invertCircle(I, k, C2) # should be C1
```


