---
title: "Jack polynomials with symbolic parameter"
author: "Stéphane Laurent"
date: '2024-04-16'
tags: R, maths, haskell, julia, Rcpp
rbloggers: yes
output:
  md_document:
    variant: markdown
    preserve_yaml: true
  html_document:
    highlight: kate
    keep_md: no
highlighter: pandoc-solarized
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, message = FALSE)
```

Nice achievment: I have been able to implement the Jack polynomials with a 
symbolic Jack parameter, in Haskell and in R. My Julia package 
**JackPolynomials**, a couple of years old, allows to get them as well:
```
julia> using JackPolynomials
julia> JackPolynomial(2, [3; 1])
(2*alpha^2 + 4*alpha + 2)*x_1^3*x_2 + (4*alpha + 4)*x_1^2*x_2^2 + (2*alpha^2 + 4*alpha + 2)*x_1*x_2^3
```

But the Haskell implementation and the R implementation required much more 
work. As compared to Julia, I also have to implement some stuff to deal 
with polynomials with symbolic coefficients. Such stuff is available in 
Julia, I didn't have to implement it myself.

In Haskell, the multivariate polynomials with symbolic coefficients are 
implemented in my package **hspray**. In R, they are implemented in my 
package **symbolicQspray** (I'm tempted to rename it to **parametricQspray**). 

Let's compute a Jack polynomial in R:
```{r}
library(jack)
JackSymPol(2, c(3, 1))
```

This is a `symbolicQspray` object. The **symbolicQspray** 
package is loaded when you load the **jack** package. A `symbolicQspray` 
object represents a multivariate polynomial whose coefficients are 
*fractions of polynomials with rational coefficients*. However I do not consider 
them as polynomials on the field of fractions of polynomials, but rather 
as polynomials with rational coefficients depending on some parameters, 
the dependence being described by a fraction of polynomials. The 
variables of the fractions of polynomials represent these parameters.

Jack polynomials have one parameter, the *Jack parameter*. It is denoted by 
`a` in the above output of `JackSymPol`.
You do not see any fraction of polynomials in this output, only polynomials. 
In fact, from the definition of the Jack polynomials, as well as from their 
implementation, the coefficients of 
these polynomials *are* fractions of polynomials in the Jack parameter. 
But you will never see a fraction in the output of `JackSymPol`: by an 
amazing theorem 
(the [Knop & Sahi formula](https://en.wikipedia.org/wiki/Jack_function#Combinatorial_formula)), 
the coefficients always are polynomials in the Jack parameter.

That explains why there are two levels of enclosing braces around the 
coefficients: `{ [ ... ] }`: the right brackets `[` and `]` enclose the
numerator of the fraction (the denominator is dropped since it is one), and 
the curly braces `{` and `}` enclose the fraction. I don't want to remove 
the right brackets when there's no denominator, because they indicate that 
the present object conceptually is a fraction of polynomials, not a 
polynomial. With the wording of my packages, this is a `ratioOfQsprays`
object, not a `qspray` object.

The `ratioOfQsprays` objects are defined in my package **ratioOfQsprays**. 
A `ratioOfQsprays` object is nothing but a pair of `qspray` objects (defined 
in my package **qspray**), the 
numerator and the denominator, but the result of an arithmetic operation
performed on these objects is always written as an irreducible fraction. 
This irreducible fraction is obtained thanks to the C++ library **CGAL**, 
which provides a very fast implementation of the greatest common divisor 
and of the division of two multivariate polynomials.

In my Haskell package **hspray**, the "parametric polynomials" are the 
objects of type `SymbolicSpray` (that I will possibly rename to 
`ParametricSpray`). These objects represent multivariate polynomials whose 
coefficients are fractions of *univariate* polynomials, whereas the R 
objects `ratioOfQsprays` represent fractions of *multivariate* polynomials. 
Univariate fractions of polynomials are enough for the Jack polynomials. 
I restricted myself to univariate polynomials because it is possible to deal 
with ratios of such polynomials with the **numeric-prelude** library, and 
I decided to use this stuff to have less work to achieve. But I have 
everything needed to introduce the fractions of multivariate polynomials 
and to use them as coefficients. Maybe in the future. Please let me know 
if you have a use case.

By the way, I'm wondering who uses my R package **jack**. I can see 
[it is downloaded](https://hadley.shinyapps.io/cran-downloads/). 
Implementing the Jack polynomials was very interesting, but I do not 
know what they are useful for. Same question for my Python package 
[**jackpy**](https://github.com/stla/jackpy), which has a couple of stars on
Github. Note that Jack polynomials with symbolic Jack parameter are not 
available in this package. This should be probably doable with a moderate 
modification of my code.

Note that Jack polynomials are often very long. For example, but this one 
is not so long:
```{r}
( jp <- JackSymPol(3, c(3, 1)) )
```

But these polynomials are symmetric, so you can get a considerably shorter 
expression by writing them as a linear combination of the monomial 
symmetric polynomials:

```{r}
compactSymmetricQspray(jp)
```

Finally, a few words about the efficiency. Here is a benchmark in R, for the 
Jack polynomial of the integer partition $[4, 2, 2, 1]$ with $5$ variables 
(everything is implemented in C++):
```{r}
library(microbenchmark)
microbenchmark(
  JackSymPol = JackSymPol(n, lambda),
  setup = {
    n <- 5
    lambda <- c(4, 2, 2, 1)
  },
  times = 5,
  unit = "seconds"
)
```

Haskell is faster: it takes about $550$ milliseconds. And unsurprisingly, 
Julia is the winner: about $350$ milliseconds (with Julia 1.9.1). 
So the **Rcpp** implementation is a bit slow.
