---
title: "Kantorovich distance with the 'ompr' package"
author: "Stéphane Laurent"
date: '2022-09-12'
tags: R, maths
rbloggers: yes
output:
  md_document:
    variant: markdown
    preserve_yaml: true
  html_document:
    highlight: kate
    keep_md: no
highlighter: pandoc-solarized
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

Do you know 
[my former blog](http://stla.github.io/stlapblog/index.html)? 
It contains four posts about the computation of the Kantorovich distance:

- [Using Julia to compute the Kantorovich distance](https://stla.github.io/stlapblog/posts/KantorovichWithJulia.html)

- [Using R to compute the Kantorovich distance](http://stla.github.io/stlapblog/posts/KantorovichWithR.html)

- [Using Scala to compute the Kantorovich distance](http://stla.github.io/stlapblog/posts/KantorovichWithScala.html)

- [The 'kantorovich' package for R](http://stla.github.io/stlapblog/posts/kantorovich-package.html)

The Julia way using the **JumP** library has the most convenient syntax:

```julia
using JuMP 
mu = [1/7, 2/7, 4/7]
nu = [1/4, 1/4, 1/2]
n = length(mu)
m = Model()
@defVar(m, p[1:n, 1:n] >= 0)
@setObjective(m, Min, sum{p[i, j], i in 1:n, j in 1:n; i != j})
for k in 1:n
  @addConstraint(m, sum(p[k, :]) == mu[k])
  @addConstraint(m, sum(p[:, k]) == nu[k])
end
solve(m)
```

This allows to get the Kantorovich distance between the two probabilities 
`mu` and `nu` corresponding to the 0-1 distance (assuming `mu` and `nu` have 
the same support). This is totally useless because one can straightforwardly 
get this distance: it is one minus the total weight of the infimum measure 
of the two probability measures (`1 - sum(pmin(mu, nu))` in R). But this is 
just for a simple illustration purpose. This problem is not trivial for another 
distance on the support of `mu` and `nu`. Encoding this distance as a 
matrix `D`, the linear programming model allowing to get the corresponding 
Kantorovich distance is obtained by replacing

```julia
sum{p[i, j], i in 1:n, j in 1:n; i != j}
```

with

```julia
sum{p[i, j] * D[i, j], i in 1:n, j in 1:n; i != j}
```

Now I want to show again how to compute the Kantorovich distance with R, but 
using another package I discovered yesterday: the **ompr** package. 
It allows to write the model with a convenient syntax, close to the 
mathematical language, similar to the one above with **JumP**. 
Here is the model:


```{r model}
library(ompr)
library(ompr.roi)
library(ROI.plugin.glpk)

mu <- c(1/7, 2/7, 4/7)
nu <- c(1/4, 1/4, 1/2)
n <- length(mu)

model <- MIPModel() |>
  add_variable(p[i, j], i = 1:n, j = 1:n, type = "continuous") |>
  add_constraint(p[i, j] >= 0, i = 1:n, j = 1:n) |>
  add_constraint(sum_over(p[i, j], j = 1:n) == mu[i], i = 1:n) |>
  add_constraint(sum_over(p[i, j], i = 1:n) == nu[j], j = 1:n) |>
  set_objective(sum_over(p[i, j], i = 1:n, j = 1:n, i != j), "min") 
```

This is nicely readable. Now we solve the problem:

```{r optimization}
optimization <- model |>
  solve_model(with_ROI(solver = "glpk"))
```

And we get the Kantorovich distance:

```{r objectiveValue, collapse=TRUE}
objective_value(optimization)
```

