---
title: "The Owen functions in quadruple precision"
author: "Stéphane Laurent"
date: "2017-12-XX"
output:
  html_document:
    keep_md: no
    toc: yes
  md_document:
    toc: yes
    variant: markdown
    preserve_yaml: true
prettify: yes
linenums: yes
prettifycss: twitter-bootstrap
tags: R, maths, statistics
highlighter: kate
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, collapse=TRUE, fig.path="./figures/OwenQuad-")
```

```{r}
dll <- "Owen128.dll"
dyn.load(dll)
.C("HsStart")
powen4qp <- function(nu, t1, t2, delta1, delta2){
  .C("owenCDF4export", as.integer(nu), as.double(t1), as.double(t2), 
     as.double(delta1), as.double(delta2), length(delta1), 
     result = numeric(length(delta1)))$result
}
```

```{r}
library(OwenQ)
```

```{r}
powerTOST <- function(alpha, delta0, Delta, sigma, n1, n2, qp = FALSE) {
  se <- sqrt(1/n1 + 1/n2) * sigma
  delta1 <- (delta0 + Delta) / se
  delta2 <- (delta0 - Delta) / se
  dof <- n1 + n2 - 2
  q <- qt(1 - alpha, dof)
  if(qp){
    powen4qp(dof, q, -q, delta1, delta2)
  }else{
    #powen4(dof, q, -q, delta1, delta2)
    R <- 0.5*sqrt(dof)*(delta1 - delta2)/q
    OwenQ1(dof, -q, delta2, R, algo=2) - OwenQ1(dof, q, delta1, R, algo=2)
  }
}
```

### Failures of `powen4`

- For $n_1=n_2=1000$, $\alpha=0.05$, $\delta=0$, $\Delta=5$, the `powen4` function abnormally takes some negative values:

```{r, echo=FALSE, fig.width=8, fig.height=4}
par(mar=c(4,4,0.4,0.4))
layout(t(c(1,2)))
sigma <- seq(65,69,len=100)
n1 <- 1000; n2 <- 1000
plot(sigma, powerTOST(0.05, 0, 5, sigma, n1, n2), type="l", lwd=2, 
     xlab=expression(sigma), ylab="power")
lines(sigma, powerTOST(0.05, 0, 5, sigma, n1, n2, qp=TRUE), col="blue", lwd=2)
abline(h=0, col="red", lty=2)
legend("topright", c("double", "quadruple"), lty=c(1,1), col=c("black", "blue"))
sigma <- seq(15,69,len=100)
plot(sigma, powerTOST(0.05, 0, 5, sigma, n1, n2), type="l", lwd=2, 
     xlab=expression(sigma), ylab="power")
lines(sigma, powerTOST(0.05, 0, 5, sigma, n1, n2, qp=TRUE), col="blue", lwd=2)
abline(h=0, col="red", lty=2)
legend("topright", c("double", "quadruple"), lty=c(1,1), col=c("black", "blue"))
```

Note that the discrepancy between `powen4` and `ipowen4` occurs only for $\sigma > 65$. 

- For $n_1=n_2=720$, $\alpha=0.05$, $\delta=0$, $\Delta=5$, we observe a discrepancy between `powen4` and `ipowen4`: 

```{r, echo=FALSE, fig.width=4, fig.height=4}
par(mar=c(4,4,0.4,0.4))
n1 <- n2 <- 720
sigma <- seq(56,57,len=100)
plot(sigma, powerTOST(0.05, 0, 5, sigma, n1, n2), type="l", lwd=2, 
     xlab=expression(sigma), ylab="power")
lines(sigma, powerTOST(0.05, 0, 5, sigma, n1, n2, qp=TRUE), col="blue", lwd=2)
legend("topright", c("double", "quadruple"), lty=c(1,1), col=c("black", "blue"))
```

The irregularity of `powen4` suggests that it returns wrong values.

- For $n_1=n_2=700$, $\alpha=0.01$, $\delta=0$, $\Delta=5$, we still observe a small discrepancy between `powen4` and `ipowen4` and an irregularity of `powen4` (on the second figure below):

```{r, echo=FALSE, fig.width=8, fig.height=4}
par(mar=c(4, 4, 0.2, 0.2))
layout(t(c(1,2)))
n1 <- n2 <- 700
sigma <- seq(35,45,len=100)
plot(sigma, powerTOST(0.01, 1, 5, sigma, n1, n2), type="l", lwd=2, 
     xlab=expression(sigma), ylab="power")
lines(sigma, powerTOST(0.01, 1, 5, sigma, n1, n2, qp=TRUE), col="blue", lwd=2)
legend("topright", c("double", "quadruple"), lty=c(1,1), col=c("black", "blue"))
n1 <- n2 <- 700
sigma <- seq(38.5,39,len=100)
plot(sigma, powerTOST(0.01, 1, 5, sigma, n1, n2), type="l", lwd=2, 
     xlab=expression(sigma), ylab="power")
lines(sigma, powerTOST(0.01, 1, 5, sigma, n1, n2, qp=TRUE), col="blue", lwd=2)
legend("topright", c("double", "quadruple"), lty=c(1,1), col=c("black", "blue"))
```

## Simulations

```{r}
nsims <- 100000
nu <- 100000
delta1 <- 4; delta2 <- 3
D <- sqrt(rchisq(nsims, df=nu)/nu)
X <- rnorm(nsims)
T1 <- (X+delta1)/D
T2 <- (X+delta2)/D
plot(T1,T2)
abline(0,1)
t1 <- 4.5; t2 <- 3
abline(v=t1, col="red", lty=2)
abline(h=t2, col="red", lty=2)
mean(T1>t1 & T2 < t2)
powen4(nu, t1, t2, delta1, delta2)
powen4qp(nu, t1, t2, delta1, delta2)
```

ça marche avec l'algo2 même en double !

```{r}
dyn.unload(dll)
```


