---
title: "An application of the resultant"
author: "Stéphane Laurent"
date: '2024-03-15'
tags: R, maths
rbloggers: yes
output:
  md_document:
    variant: markdown
    preserve_yaml: true
  html_document:
    highlight: kate
    keep_md: no
highlighter: pandoc-solarized
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.path = "./figures/resultant-algebraicCurves-", 
                      fig.width = 3.5, fig.align = "center",
                      collapse = TRUE, warning = FALSE)
```

I will soon release an update of **qspray** on CRAN as well as a new package: 
**resultant**.
This post shows an application of these two packages.

Consider the two algebraic curves $f(x,y)=0$ and $g(x,y)=0$ where 
$$
f(x, y) = y^4 - y^3 + y^2 - 2x^2y + x^4 
\quad \text{ and } \quad
g(x, y) = y - 2x^2.
$$
We will derive their intersection points. First of all, let's plot them.

```{r}
f <- function(x, y) {
  y^4 - y^3 + y^2 - 2*x^2*y + x^4
}

g <- function(x, y) {
  y - 2*x^2
}

# contour line for f(x,y)=0
x <- seq(-1.2, 1.2, len = 2000)
y <- seq(0, 1, len = 2000)
z <- outer(x, y, f)
crf <- contourLines(x, y, z, levels = 0)
# contour line for g(x,y)=0
x <- seq(-1, 1, len = 2000)
y <- seq(0, 1.5, len = 2000)
z <- outer(x, y, g)
crg <- contourLines(x, y, z, levels = 0)
```

Theoretically there is only one contour line for both. But for some technical 
reasons, `crf` is split into several pieces:

```{r}
length(crf)
```

So we need a helper function to construct the dataframe that we will pass to 
`ggplot2::geom_path`:

```{r}
intercalateNA <- function(xs) {
  if(length(xs) == 1L) {
    xs[[1L]]
  } else {
    c(xs[[1L]], NA, intercalateNA(xs[-1L]))
  }
}
contourData <- function(cr) {
  xs <- lapply(cr, `[[`, "x")
  ys <- lapply(cr, `[[`, "y")
  data.frame("x" = intercalateNA(xs), "y" = intercalateNA(ys))
}

datf <- contourData(crf)
datg <- contourData(crg)
```

I also plot the intersection points that we will derive later:

```{r ggplot, fig.height=3, message=FALSE}
datPoints <- data.frame("x" = c(-0.5, 0, 0.5), "y" = c(0.5, 0, 0.5))

library(ggplot2)
ggplot() +
  geom_path(aes(x, y), data = datf, linewidth = 1, color = "blue") +
  geom_path(aes(x, y), data = datg, linewidth = 1, color = "green") +
  geom_point(aes(x, y), data = datPoints, size = 2)
```

Now we compute the *resultant* of the two polynomials with respect to $x$:

```{r}
# define the two polynomials
library(qspray)
x <- qlone(1)
y <- qlone(2)
P <- f(x, y)
Q <- g(x, y)
# compute their resultant with respect to x
Rx <- resultant::resultant(P, Q, var = 1) # var=1 <=> var="x"
prettyQspray(Rx, vars = "x")
```

We need the roots of the resultant $R_x$. I use **giacR** to get them:

```{r}
library(giacR)
giac <- Giac$new()
command <- sprintf("roots(%s)", prettyQspray(Rx, vars = "x"))
giac$execute(command)
giac$close()
```

Thus there are two roots: $0$ and $1/2$ (the output of **GIAC** also provides 
their multiplicities). 
Luckily, they are rational, so we can use `substituteQspray` to replace $y$ 
with each of these roots in $P$ and $Q$. We firstly do the substitution $y=0$:

```{r}
Px <- substituteQspray(P, c(NA, "0"))
Qx <- substituteQspray(Q, c(NA, "0"))
prettyQspray(Px, "x")
prettyQspray(Qx, "x")
```
Clearly, $0$ is the unique common root of these two polynomials. 
One can conclude that $(0,0)$ is an intersection point.

Now we do the substitution $y=1/2$:
```{r}
Px <- substituteQspray(P, c(NA, "1/2"))
Qx <- substituteQspray(Q, c(NA, "1/2"))
prettyQspray(Px, "x")
prettyQspray(Qx, "x")
```
It is easy to see that $x= \pm 1/2$ are the roots of the second polynomial. 
And one can check that they are also some roots of the first one. 
One can conclude that $(-1/2, 1/2)$ and $(1/2, 1/2)$ are some intersection 
points. 

And we're done.

