---
title: "A R list to a Haskell list, with the `clipr` package"
author: "Stéphane Laurent"
date: "2018-03-14"
output:
  md_document:
    variant: markdown
    preserve_yaml: true
  html_document:
    keep_md: no
prettify: yes
linenums: yes
prettifycss: twitter-bootstrap
tags: R, haskell
highlighter: kate
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```


In the [previous post](https://laustep.github.io/stlahblog/posts/Rmatrix2HaskellList.html), I have shown how to convert a R matrix to a Haskell list. 
Now I will show how to convert a R list to a Haskell list. 


Let's take this list for illustration:

```{r}
L <- list(c(1,2,3), c(1,2,3,4), c(1,2))
L
```


So, for Haskell, you want:

```haskell
[ [1, 2, 3]
 ,[1, 2, 3, 4]
 ,[1, 2] 
]
```

Again, the `clipr` package is your friend. 
First, write the matrix in the clipboard, like this:

```{r}
library(clipr)
write_clip(L, breaks="],\n", sep=", ")
```

No we will deal with `gsub`, `sub` and finally `cat` to write the output to a file.

```{r}
myfile <- "list.txt"
cat("[",
    sub("\\)", "]\n]", gsub("\\)]", "]", gsub("c\\(", "[" ,read_clip()))), 
    sep="\n", file=myfile)
```

And then, here is the content of `list.txt`: 

```txt
[
[1, 2, 3],
[1, 2, 3, 4],
[1, 2]
]
```

Now copy-paste to Haskell, it is ready.


### Update 2018-04-18

I've realized there is an issue in the case when some elements of the list are vectors of length one. It is fixed in the following function:

```{r}
list2list <- function(L, outfile="list.txt"){
  singletons_idxs <- which(lapply(L,length)==1)
  singletons <- L[singletons_idxs]
  L <- replace(L, singletons_idxs, sprintf("c(%s)", singletons))
  write_clip(L, breaks="],\n", sep=", ")
  cat("[",
      sub("\\)", "]\n]", gsub("\\)]", "]", gsub("c\\(", "[" ,read_clip()))), 
      sep="\n", file=outfile)
}
```

Let's test it. 

```{r}
L <- list(c(1,2,3), 0, c(1,2,3,4), c(1,2))
list2list(L)
```

Here is the content of `list.txt`: 

```txt
[
[1, 2, 3],
[0],
[1, 2, 3, 4],
[1, 2]
]
```

Fine.