---
title: "A R matrix to a Haskell list, with the `clipr` package"
author: "Stéphane Laurent"
date: "2018-03-13"
output:
  md_document:
    variant: markdown
    preserve_yaml: true
  html_document:
    keep_md: no
prettify: yes
linenums: yes
prettifycss: twitter-bootstrap
tags: R, haskell
highlighter: kate
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```



Assume you have a matrix in R, and you want to use it in Haskell as a list. Here is a way to go. 


I will take a small matrix for the illustration.

```{r}
M <- rbind(
  c(1,2,1.5),
  c(0.5,2,3),
  c(5,4.3,7)
)
M
```


So, for Haskell, you want:

```haskell
[ [1.0,  2.0,  1.5]
 ,[0.5,  2.0,  3.0]
 ,[5.0,  4.3,  7.0] ]
```

It would be very painful to do it by hands. But the `clipr` package is your friend. 
First, write the matrix in the clipboard, like this:

```{r}
library(clipr)
write_clip(M, breaks="],\n", sep=", ")
```

No we will use `cat` to write the output to a file.

```{r}
library(magrittr)
myfile <- "matrix.txt"
paste0("[", read_clip()) %>% 
  cat(sep="\n", file=myfile) %>% 
    cat("]", sep="", file=myfile, append=TRUE)
```

And then, here is the content of `matrix.txt`: 

```txt
[1, 2, 1.5],
[0.5, 2, 3],
[5, 4.3, 7

]
```

Well, not totally perfect. But now it's a child game to complete the output before copying it to Haskell:


```txt
[
[1, 2, 1.5],
[0.5, 2, 3],
[5, 4.3, 7]
]
```
